### 2016-11-03 paper.tex
Add derivation steps
At the moment this is little more than a formula dump.

\documentclass[english,listof=totoc]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}

\title{Estimation of Matrix-t distributions with Expectation
Maximization}

\author{Zahari Kassabov\thanks{TIF-UNIMI-2016-5}\\
        Dipartimento di Fisica, Universit\`a di Torino and INFN, Sezione di Torino\\
		TIF Lab, Dipartimento di Fisica, Universit\`a di Milano\\
        E-mail: \email{kassabov@to.infn.it}}

\begin{document}

\begin{abstract}
We derive the expectation-maximization (EM) formulae for
Matrix-$t$ distributions and discuss their practical implementation.
\end{abstract}

\paragraph{Introduction}
Matrix-$t$ distributions have been applied to several predictive
problems, such as spatial interpolation applied to the prediction of
pollution concentration, or recommendation systems \cite{NIPS2007_3203}???
...

\paragraph{Derivation}

Most of the results are based on the integral

\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}dS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}\label{eq:intmultgammadef}
\end{equation}
where we integrate $S$ over the space of positive definite $p\times p$
matrices, $A$ is a $p\times p$ nonsingular matrix, and the multivariate
gamma function $\Gamma_{p}(a)$ is given in terms of the standard
gamma function by:
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+(1-j)/2\right)\label{eq:multgammadef}
\end{equation}
we also define the multivariate digamma function
\[
\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi(a+(1-j)/2)
\]

The PDF of a random matrix $p\times m$ $T$ following a matrix-t
distribution
\[
T_{p,m}(M,\Sigma,\Omega,n)
\]
is given by:
\begin{equation}
\begin{split}\textrm{tpdf}(T,M,\Sigma,\Omega,n)=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}(\det\Omega)^{\frac{p}{2}}(\det\Sigma)^{\frac{m}{2}}}\times\\
\det\left(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right){}^{\frac{1-m-p-n}{2}}
\end{split}
\label{eq:matrixtpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\Omega$
is a $m\times m$ positive definite matrix, $M$ is a $p\times m$
matrix, and $n$ is the number of degrees of freedom.

Following {[}book{]}, we begin by deriving the PDF of the matrix-t,
starting from a matrix normal distributed random matrix $X$, where
row-covariance matrix $S$ is distributed following a Wishart distribution
with the appropriate count of degrees of freedom (independent from
$X$), and marginalizing over $S$. We repeat the results from {[}book{]}
because, along the way, we will write several results that will be
useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X$ follows the matrix-normal distribution

\begin{equation}
N_{p,m}(M,S,\Omega)\label{eq:ndist}
\end{equation}
if its PDF is given by:
\begin{equation}
\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
where $S$ is a $p\times p$ positive definite matrix (the covariance
of the rows of $X$), $\Omega$ is a $m\times m$ positive definite
matrix (the covariance of the columns of $X$) and $M$ is a $p\times m$
matrix (the mean of $X$).

In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
\begin{equation}
W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
\end{equation}
 if its PDF is given by:
\begin{equation}
\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
is the degrees of freedom and $\Gamma_{p}$ is the multivariate gamma
function defined in \ref{eq:multgammadef}. We will show that the
random matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
a matrix-$t$ distribution with $n=\nu-p+1$. The Jacobian of this
transformation is 
\begin{equation}
(\det S)^{\frac{1}{2}m}.\label{eq:jac}
\end{equation}

The joint PDF of $T$ and $S$ is the product of Eq.~\ref{eq:ndistpdf},
Eq.~\ref{eq:wdistpdf} and Eq.~\ref{eq:jac}:
\begin{equation}
\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right)
\end{split}
\label{eq:jointpdf}
\end{equation}
Finally we obtain the PDF of the matrix-$t$ distribution Eq.~\ref{eq:matrixtpdf}
by integrating over $S$ in the above expression, using directly Eq.~\ref{eq:intmultgammadef}
and slightly rearranging the terms (in particular, noting that $\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).

\subsection{Calculating the Expectation-maximization steps}

The expectation maximization formalism entails calculating 
\begin{equation}
E=\sum_{i}^{N}\frac{\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})dS}{\textrm{tpdf}(T,M,\Sigma,\Omega,n)}\label{eq:expdef}
\end{equation}
where $\{T_{1}...T_{N}\}$ is an input set of data from which we wish
to estimate the maximum likelihood parameters of the distribution.
We omit writing explicitly the dependence of $E$ on all the parameters,
i.e. $E\equiv E(T_{i},S,M,\Sigma,\Omega,n,M_{t},\Sigma_{t},\Omega_{t},n_{t})$.
The values of the new guesses are found by computing the maximum with
respect to the parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$.

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\[
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\]
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix-$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In obtain to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve \ref{eq:firstlogpart} by substituting this result directly.

In order to compute the second nontrivial log term,
\[
\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))dS
\]
we note that, due to the linearity of the trace, the result is equivalent
to solving a matrix valued integral
\[
\textrm{tr}\left[\left(\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)SdS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
\]
The enclosed integral also follows from Eq.~\ref{eq:intmultgammadef},
after taking the matrix derivative with respect to $A$. Using, for
symmetric $S$
\begin{equation}
\frac{d}{dA}\det A=(\det A)A^{-1}\label{eq:detderivative}
\end{equation}
and
\[
\frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
\]
we have 
\[
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}SdS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
\]
from where again, we can obtain the result by direct substitution.

Combining results of the two integrals together, and with adequate
rearrangements we obtain the integrated form of Eq.~\ref{eq:expdef}:
\begin{equation}
\begin{split}E=\sum_{i}^{N}n_{t}\psi_{p}\left(\frac{1}{2}(m+n+p-1)\right)-2\log\Gamma_{p}\left(\frac{1}{2}(n_{t}+p-1)\right)+p\log\det\Omega_{t}^{-1}\\
-n_{t}\log\det\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)+(n_{t}+p-1)\log\det\Sigma_{t}\\
-(m+n+p-1)\textrm{tr}\left(\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left((T_{i}-M_{t})\Omega_{t}^{-1}(T_{i}-M_{t})^{\textrm{t}}+\Sigma_{t}\right)\right)
\end{split}
\label{eq:fsimp}
\end{equation}
The maximization formulas are obtained setting the derivatives of
this equation to zero.

For $M_{t}$ we have:
\[
\frac{dE}{dM_{t}}=0\Rightarrow\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left(T_{i}-M_{t}\right)\Omega_{t}^{-1}=0
\]
and because $\Omega_{t}$ is nonsingular, we can obtain $M_{t}$ as
a weighted average of the input data:
\[
M_{t}=\frac{1}{N}\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}T_{i}=\frac{1}{N}\sum_{i}^{N}W_{i}T_{i}
\]
where we have defined the weight matrices $W_{i}$ as
\[
W_{i}=\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}
\]

$\Sigma_{t}$ is given in terms of the inverse of the mean weight
by:
\[
\Sigma_{t}=\frac{n_{t}+p-1}{n+m+p-1}\left[\frac{1}{N}\sum_{i}^{N}W_{i}\right]^{-1}
\]
Note that it depends on $n_{t}$.

For $\Omega_{t}$ we obtain:
\[
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\]
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated we have:
\[
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\[
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\]

\bibliography{references}
\end{document}

### 2016-11-23 paper.tex
Added author info and some minor changes.

@@
 
-\title{Estimation of Matrix-t distributions with Expectation
-Maximization}
+\title{ML Estimation of Matrix t Distributions with Expectation Maximization}
 
@@
 		TIF Lab, Dipartimento di Fisica, Universit\`a di Milano\\
-        E-mail: \email{kassabov@to.infn.it}}
+        E-mail: \email{kassabov@to.infn.it}\\
+        \\
+        Hsien-Ching Kao\\
+        Wolfram Research, Inc.\\
+		Champaign, Illinois 61820, USA\\
+		E-mail: \email{sp000088@gmail.com}
+		}
 
@@
 
+\maketitle
+
 \begin{abstract}
-We derive the expectation-maximization (EM) formulae for
-Matrix-$t$ distributions and discuss their practical implementation.
+Matrix-$t$ distribution has applications in various areas of applied statistics, including recommendation systems, remote sensing, etc.
+In this article we provide an expectation-maximization (EM) formulae for Matrix-$t$ distributions and discuss their practical implementation.
 \end{abstract}
@@
 \paragraph{Introduction}
-Matrix-$t$ distributions have been applied to several predictive
-problems, such as spatial interpolation applied to the prediction of
-pollution concentration, or recommendation systems \cite{NIPS2007_3203}???
+Matrix-$t$ distributions have been applied to several predictive problems, such as spatial interpolation applied to the prediction of pollution concentration \cite{}, recommendation systems \cite{NIPS2007_3203},
 ...
@@
 \begin{equation}
-\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}dS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}\label{eq:intmultgammadef}
+\int_{S\succ 0}e^{-\textrm{tr}SA}(\det S)^{x}dS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}\label{eq:intmultgammadef}
 \end{equation}

### 2016-12-01 paper.tex
Minor adjustments.

@@
 \begin{abstract}
-Matrix-$t$ distribution has applications in various areas of applied statistics, including recommendation systems, remote sensing, etc.
-In this article we provide an expectation-maximization (EM) formulae for Matrix-$t$ distributions and discuss their practical implementation.
+Matrix--$t$ distribution has applications in various areas of applied statistics, including recommendation systems, remote sensing, etc.
+In this article we provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
 \end{abstract}
@@
 \paragraph{Introduction}
-Matrix-$t$ distributions have been applied to several predictive problems, such as spatial interpolation applied to the prediction of pollution concentration \cite{}, recommendation systems \cite{NIPS2007_3203},
+Matrix--$t$ distributions have been applied to several predictive problems, such as spatial interpolation applied to the prediction of pollution concentration \cite{}, recommendation systems \cite{NIPS2007_3203}, and
 ...
@@
 
-Most of the results are based on the integral
+The derivation of EM formulae for matrix--$t$ distributions is mainly based on the following integral identity:
 
 \begin{equation}
-\int_{S\succ 0}e^{-\textrm{tr}SA}(\det S)^{x}dS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}\label{eq:intmultgammadef}
+\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
 \end{equation}
-where we integrate $S$ over the space of positive definite $p\times p$
-matrices, $A$ is a $p\times p$ nonsingular matrix, and the multivariate
-gamma function $\Gamma_{p}(a)$ is given in terms of the standard
-gamma function by:
+where the domain of integration of $S$ is the set of real symmetric positive definite
+matrices with dimension $p$, $A$ is a $p\times p$ nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma function given in terms of the standard gamma function $\Gamma(\cdot)$ by
 \begin{equation}
-\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+(1-j)/2\right)\label{eq:multgammadef}
+\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
 \end{equation}
-we also define the multivariate digamma function
-\[
-\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi(a+(1-j)/2)
-\]
+For notation convenience, the matrix determinant is represented as $|\cdot|$. We also define the multivariate digamma function
+\begin{equation}
+\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
+\end{equation}
 
-The PDF of a random matrix $p\times m$ $T$ following a matrix-t
-distribution
-\[
-T_{p,m}(M,\Sigma,\Omega,n)
-\]
-is given by:
+A random $p\times m$ matrix $T$ following a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$
+has probability density
 \begin{equation}
-\begin{split}\textrm{tpdf}(T,M,\Sigma,\Omega,n)=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}(\det\Omega)^{\frac{p}{2}}(\det\Sigma)^{\frac{m}{2}}}\times\\
-\det\left(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right){}^{\frac{1-m-p-n}{2}}
-\end{split}
+f(T;\left\{M,\Sigma,\Omega,n\right\})=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
+|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-n}{2}},
 \label{eq:matrixtpdf}
 \end{equation}
-where $\Sigma$ is a $p\times p$ positive definite matrix, $\Omega$
-is a $m\times m$ positive definite matrix, $M$ is a $p\times m$
-matrix, and $n$ is the number of degrees of freedom.
+where $\Sigma$ and $\Omega$ are both real symmetric positive definite matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$ matrix indicates the offset, and $n$ is the number of degrees of freedom.
 
-Following {[}book{]}, we begin by deriving the PDF of the matrix-t,
-starting from a matrix normal distributed random matrix $X$, where
-row-covariance matrix $S$ is distributed following a Wishart distribution
-with the appropriate count of degrees of freedom (independent from
-$X$), and marginalizing over $S$. We repeat the results from {[}book{]}
-because, along the way, we will write several results that will be
-useful in the derivation of the EM formulas.
+Following {[}book{]} we begin by deriving the probability density of matrix--$t$ distribution starting from a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$ with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale parameter $S$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$.
+Note that the row scale parameter $S$ is a real symmetric positive definite matrix.
+We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.
 
-The $p\times m$ random matrix $X$ follows the matrix-normal distribution
-
-\begin{equation}
-N_{p,m}(M,S,\Omega)\label{eq:ndist}
-\end{equation}
-if its PDF is given by:
+The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has probability density
 \begin{equation}
@@
 where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
-a matrix-$t$ distribution with $n=\nu-p+1$. The Jacobian of this
+a matrix--$t$ distribution with $n=\nu-p+1$. The Jacobian of this
 transformation is 
@@
 \end{equation}
-Finally we obtain the PDF of the matrix-$t$ distribution Eq.~\ref{eq:matrixtpdf}
+Finally we obtain the PDF of the matrix--$t$ distribution Eq.~\ref{eq:matrixtpdf}
 by integrating over $S$ in the above expression, using directly Eq.~\ref{eq:intmultgammadef}
@@
 compute the corresponding integral, while for the rest, the integral
-is proportional to the PDF of the matrix-$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
+is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
 In obtain to compute:

### 2016-12-11 paper.tex
Fix header
Now compiles

@@
 		TIF Lab, Dipartimento di Fisica, Universit\`a di Milano\\
-        E-mail: \email{kassabov@to.infn.it}\\
-        \\
+        \texttt{kassabov@to.infn.it}\\
+        \and
         Hsien-Ching Kao\\
@@
 		Champaign, Illinois 61820, USA\\
-		E-mail: \email{sp000088@gmail.com}
+		\texttt{sp000088@gmail.com}
 		}

### 2016-12-11 paper.tex
Change abstract
Put what we do on top.

@@
 \begin{abstract}
-Matrix--$t$ distribution has applications in various areas of applied statistics, including recommendation systems, remote sensing, etc.
-In this article we provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
+We provide expectation-maximization (EM) formulae for the maximum
+likelihood estimation of matrix--$t$ distributions and discuss their
+practical implementation.  The result can be applied to problems where
+the parameters of a matrix--$t$ distribution need to be obtained from
+data, such as the design of recommendation systems, or remote sensing
+algorithms.
 \end{abstract}

### 2016-12-11 paper.tex
Fix citations and add one reference

@@
 \usepackage{amsthm}
-\usepackage{mathtools}
+\usepackage[authoryear]{natbib}
+\usepackage[unicode=true,pdfusetitle,
+ bookmarks=true,bookmarksnumbered=false,bookmarksopen=false,
+ breaklinks=false,pdfborder={0 0 0},pdfborderstyle={},backref=false,colorlinks=false]
+ {hyperref}\usepackage{mathtools}
 
@@
 \paragraph{Introduction}
-Matrix--$t$ distributions have been applied to several predictive problems, such as spatial interpolation applied to the prediction of pollution concentration \cite{}, recommendation systems \cite{NIPS2007_3203}, and
-...
+Matrix--$t$ distributions have been applied to several predictive
+problems, such as spatial interpolation applied to the prediction of
+pollution concentration \citep{KIBRIA2006785}, recommendation systems
+\citep{NIPS2007_3203}, and ...
 
@@
 
+\bibliographystyle{unsrtnat}
 \bibliography{references}

### 2016-12-11 paper.tex
Improve introduction

@@
 pollution concentration \citep{KIBRIA2006785}, recommendation systems
-\citep{NIPS2007_3203}, and ...
+\citep{NIPS2007_3203}. In these situations it is necessary to obtain
+the parameters of the metrix--$t$ distribution (consisting of the
+number of degrees of freedom, a location matrix and two scale
+matrices) that describe a set of given matrix valued observations.
+While the exact formulae for Maximum Likelihood Estimation (MLE) are
+unwieldy and slow to compute, the problem can be simplified if instead
+we obtain the distribution parameters using Expectation-Maximization
+(EM). We obtain can obtain the EM expressions by viewing the
+matrix--$t$  distribution as a matrix-normal distribution where one of
+the scale matrices is sampled from a Wishart distribution.
 
@@
 
-The derivation of EM formulae for matrix--$t$ distributions is mainly based on the following integral identity:
+The derivation of the EM formulae is mainly based on the following integral identity:
 

### 2016-12-11 paper.tex
Some work on the text
Improve wording and add a reference to the book.

@@
 
-The derivation of the EM formulae is mainly based on the following integral identity:
+The derivation of the EM formulae is mainly based on the following
+integral identity:
 
@@
 \end{equation}
-where the domain of integration of $S$ is the set of real symmetric positive definite
-matrices with dimension $p$, $A$ is a $p\times p$ nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma function given in terms of the standard gamma function $\Gamma(\cdot)$ by
+where the domain of integration of $S$ is the set of real symmetric
+positive definite matrices with dimension $p$, $A$ is a $p\times p$
+nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
+function, given in terms of the standard gamma function $\Gamma(\cdot)$
+by
 \begin{equation}
@@
 \end{equation}
-For notation convenience, the matrix determinant is represented as $|\cdot|$. We also define the multivariate digamma function
-\begin{equation}
-\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
+For notation convenience, the matrix determinant is represented as
+$|\cdot|$. We also define the multivariate digamma function
+%
+\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
+	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
 \end{equation}
 
-A random $p\times m$ matrix $T$ following a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$
-has probability density
+A random $p\times m$ matrix $T$ following a matrix--$t$ distribution
+$T_{p,m}(M,\Sigma,\Omega,n)$ has the probability density
 \begin{equation}
@@
 \end{equation}
-where $\Sigma$ and $\Omega$ are both real symmetric positive definite matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$ matrix indicates the offset, and $n$ is the number of degrees of freedom.
+where $\Sigma$ and $\Omega$ are both real symmetric positive definite
+matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$
+matrix that indicates the location of the distribution, and $n$ is the
+number of degrees of freedom.
 
-Following {[}book{]} we begin by deriving the probability density of matrix--$t$ distribution starting from a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$ with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale parameter $S$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$.
-Note that the row scale parameter $S$ is a real symmetric positive definite matrix.
-We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.
+Following \citet{gupta1999matrix}, we begin by deriving the
+probability density of the matrix--$t$ distribution starting from
+a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$
+with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale
+parameter $S$ of appropriate degrees of freedom $\nu$ (independent
+from $X$), and marginalizing over $S$. Note that the row scale
+parameter $S$ is a real symmetric positive definite matrix. We repeat
+the results here because, along the way, we will utilize several
+results that are useful in the derivation of the EM formulas.
 
-The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has probability density
+The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has
+the probability density %
 \begin{equation}
-\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
+	\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det
+	S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
 \end{equation}
-where $S$ is a $p\times p$ positive definite matrix (the covariance
-of the rows of $X$), $\Omega$ is a $m\times m$ positive definite
-matrix (the covariance of the columns of $X$) and $M$ is a $p\times m$
-matrix (the mean of $X$).
+where $S$ is a $p\times p$ positive definite matrix (the covariance of
+the rows of $X$), $\Omega$ is a $m\times m$ positive definite matrix
+(the covariance of the columns of $X$) and $M$ is a $p\times m$ matrix
+(the mean of $X$).
 
@@
 where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
-is the degrees of freedom and $\Gamma_{p}$ is the multivariate gamma
-function defined in \ref{eq:multgammadef}. We will show that the
-random matrix
+is the number of degrees of freedom and $\Gamma_{p}$ is the
+multivariate gamma function defined in \ref{eq:multgammadef}. We will
+show that the random matrix
 \begin{equation}
@@
 \end{equation}
-Finally we obtain the PDF of the matrix--$t$ distribution Eq.~\ref{eq:matrixtpdf}
-by integrating over $S$ in the above expression, using directly Eq.~\ref{eq:intmultgammadef}
-and slightly rearranging the terms (in particular, noting that $\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).
+Finally we obtain the PDF of the matrix--$t$ distribution
+Eq.~\ref{eq:matrixtpdf} by integrating over $S$ in the above
+expression, using directly Eq.~\ref{eq:intmultgammadef} and slightly
+rearranging the terms (in particular, noting that
+$\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times
+p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).
 
@@
 is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
-In obtain to compute:
+In order to compute:
 \begin{equation}
@@
 \end{equation}
-We can solve \ref{eq:firstlogpart} by substituting this result directly.
+We can solve Eq.~\ref{eq:firstlogpart} by substituting this result directly.
 
@@
 and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
-that needs to be estimated we have:
+that needs to be estimated, we have:
 \[

### 2016-12-11 paper.tex
Make everything an equation

@@
 we have: 
-\[
+\begin{equation}
 \begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
@@
 \end{split}
-\]
+\end{equation}
 where the first two terms depend on $S$, and therefore we need to
@@
 In order to compute the second nontrivial log term,
-\[
+\begin{equation}
 \int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))dS
-\]
+\end{equation}
 we note that, due to the linearity of the trace, the result is equivalent
 to solving a matrix valued integral
-\[
+\begin{equation}
 \textrm{tr}\left[\left(\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)SdS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
-\]
+\end{equation}
 The enclosed integral also follows from Eq.~\ref{eq:intmultgammadef},
@@
 and
-\[
+\begin{equation}
 \frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
-\]
+\end{equation}
 we have 
-\[
+\begin{equation}
 \int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}SdS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
-\]
+\end{equation}
 from where again, we can obtain the result by direct substitution.
@@
 For $M_{t}$ we have:
-\[
+\begin{equation}
 \frac{dE}{dM_{t}}=0\Rightarrow\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left(T_{i}-M_{t}\right)\Omega_{t}^{-1}=0
-\]
+\end{equation}
 and because $\Omega_{t}$ is nonsingular, we can obtain $M_{t}$ as
 a weighted average of the input data:
-\[
+\begin{equation}
 M_{t}=\frac{1}{N}\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}T_{i}=\frac{1}{N}\sum_{i}^{N}W_{i}T_{i}
-\]
+\end{equation}
 where we have defined the weight matrices $W_{i}$ as
-\[
+\begin{equation}
 W_{i}=\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}
-\]
+\end{equation}
 
@@
 by:
-\[
+\begin{equation}
 \Sigma_{t}=\frac{n_{t}+p-1}{n+m+p-1}\left[\frac{1}{N}\sum_{i}^{N}W_{i}\right]^{-1}
-\]
+\end{equation}
 Note that it depends on $n_{t}$.
@@
 For $\Omega_{t}$ we obtain:
-\[
+\begin{equation}
 \Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
-\]
+\end{equation}
 and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
 that needs to be estimated, we have:
-\[
+\begin{equation}
 -p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
-\]
+\end{equation}
 otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
 $n_{t}$:
-\[
+\begin{equation}
 \log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
-\]
+\end{equation}
 

### 2017-01-03 paper.tex
Worked on page 2 and 3.

@@
 \end{equation}
-where the domain of integration of $S$ is the set of real symmetric
+where the domain of integration is the set of real symmetric
 positive definite matrices with dimension $p$, $A$ is a $p\times p$
@@
 probability density of the matrix--$t$ distribution starting from
-a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$
-with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale
-parameter $S$ of appropriate degrees of freedom $\nu$ (independent
-from $X$), and marginalizing over $S$. Note that the row scale
-parameter $S$ is a real symmetric positive definite matrix. We repeat
-the results here because, along the way, we will utilize several
-results that are useful in the derivation of the EM formulas.
+a matrix normal distributed random variate $X\sim N_{p,m}(M,S,\Omega)$
+with Wishart distributed row scale parameter $S \sim W_{p}\left(\Sigma,\nu\right)$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale
+parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.
 
-The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has
-the probability density %
+The $p\times m$ random matrix $X \sim N_{p,m}(M,S,\Omega)$ has density
 \begin{equation}
-	\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det
-	S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
+	\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right),\label{eq:ndistpdf}
 \end{equation}
-where $S$ is a $p\times p$ positive definite matrix (the covariance of
-the rows of $X$), $\Omega$ is a $m\times m$ positive definite matrix
-(the covariance of the columns of $X$) and $M$ is a $p\times m$ matrix
-(the mean of $X$).
+where $S$ and $\Omega$ are symmetric positive definite matrices of dimension $p$ and $m$, respectively, and $M$ (the expectation of $X$) is a $p\times m$ matrix.
+The matrix $S$ is proportional to the covariance of the rows of $X$ while the matrix $\Omega$ is proportional to the covariance of the columns of $X$.
+In turn the $p\times p$ matrix $S\sim W_{p}\left(\Sigma,\nu\right)$ has density
+\begin{equation}
+\frac{|S|^{\frac{\nu-p-1}{2}}}{2^{\nu p/2}|\Sigma|^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}\exp\left(-\textrm{tr}\left[\Sigma^{-1}S\right]\right),\label{eq:wdistpdf}
+\end{equation}
+where $\Sigma$ is a $p\times p$ symmetric positive definite matrix, $\nu>p-1$
+is the number of degrees of freedom, and $\Gamma_{p}$ is the
+multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.
 
-In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
-\begin{equation}
-W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
-\end{equation}
- if its PDF is given by:
-\begin{equation}
-\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
-\end{equation}
-where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
-is the number of degrees of freedom and $\Gamma_{p}$ is the
-multivariate gamma function defined in \ref{eq:multgammadef}. We will
-show that the random matrix
+We will show that the matrix
 \begin{equation}
@@
 \end{equation}
-where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
-a matrix--$t$ distribution with $n=\nu-p+1$. The Jacobian of this
-transformation is 
+follows a matrix--$t$ distribution with degrees of freedom $n=\nu-p+1$.
+Here $S^{\frac{1}{2}}$ is the lower Cholesky factor of $S$, i.e., $S=S^{\frac{1}{2}}(S^{\frac{1}{2}})^t$.
+The Jacobian determinant of the transformation \eqref{eq:change} is
 \begin{equation}
-(\det S)^{\frac{1}{2}m}.\label{eq:jac}
+\left|\frac{\partial T}{\partial X}\right|=|S|^{-\frac{m}{2}}.\label{eq:jac}
 \end{equation}
 
-The joint PDF of $T$ and $S$ is the product of Eq.~\ref{eq:ndistpdf},
-Eq.~\ref{eq:wdistpdf} and Eq.~\ref{eq:jac}:
+Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint PDF of $T$ and $S$ is given as
 \begin{equation}
 \begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
-\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right)
+\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right).
 \end{split}
@@
 \end{equation}
-Finally we obtain the PDF of the matrix--$t$ distribution
-Eq.~\ref{eq:matrixtpdf} by integrating over $S$ in the above
-expression, using directly Eq.~\ref{eq:intmultgammadef} and slightly
-rearranging the terms (in particular, noting that
+Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
+by integrating over $S$ in the above expression, using Eq.~\eqref{eq:intmultgammadef} and slightly
+rearranging the terms. In particular noting that
 $\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times
-p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).
+p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$.
 

### 2017-01-05 paper.tex
Formulas editing.

@@
 \begin{equation}
-\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
+\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}\,dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
 \end{equation}
@@
 \end{equation}
-
-Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint PDF of $T$ and $S$ is given as
+Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint density of $T$ and $S$ is given as
 \begin{equation}
-\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
-\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right).
+\begin{split}j(T,S;\left\{M,\Sigma,\Omega,n\right\})=\frac{\pi^{-\frac{mp}{2}}|\Sigma|^{\frac{n+p-1}{2}}|\Omega|^{\frac{n+m-2}{2}}|S|^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
+\exp\left(-\frac{1}{2}\textrm{tr}\left[S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right]\right).
 \end{split}
@@
 \end{equation}
+
 Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
-by integrating over $S$ in the above expression, using Eq.~\eqref{eq:intmultgammadef} and slightly
+by integrating over $S$ in the above expression, using Eq.~\eqref{eq:change} and slightly
 rearranging the terms. In particular noting that
-$\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times
-p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$.
+$|\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|=|\Sigma||I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|$.
 
@@
 \begin{equation}
-E=\sum_{i}^{N}\frac{\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})dS}{\textrm{tpdf}(T,M,\Sigma,\Omega,n)}\label{eq:expdef}
+E=\sum_{i=1}^{N}\frac{\int_{S\succ 0}j(T_{i},S,M,\Sigma,\Omega,n)\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\,dS}{f(T;\left\{M,\Sigma,\Omega,n\right\})},\label{eq:expdef}
 \end{equation}
-where $\{T_{1}...T_{N}\}$ is an input set of data from which we wish
-to estimate the maximum likelihood parameters of the distribution.
-We omit writing explicitly the dependence of $E$ on all the parameters,
-i.e. $E\equiv E(T_{i},S,M,\Sigma,\Omega,n,M_{t},\Sigma_{t},\Omega_{t},n_{t})$.
-The values of the new guesses are found by computing the maximum with
-respect to the parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$.
+where $\{T_i\}_{i=1}^N$ is a dataset consists of matrix elements $T_i$ from which we wish to estimate the maximum likelihood parameters of the distribution.
+We omit writing explicitly the dependence of $E$ on all the parameters.
+The values of the new guesses are found by maximizing with respect to the parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$.
 
-We can split the integral in Eq.~\ref{eq:expdef} by the terms of
-$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
-a global factor $\frac{1}{2}$ and terms that do not depend on the
-maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
-we have: 
+We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the
+maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$ we have
 \begin{equation}
-\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
--n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
+\begin{split}\log j(T_{i},S;\left\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\right\})\sim n_{t}\log|S|+\textrm{tr}\left[S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\\
+-n_{t}p\log2-p\log |\Omega_{t}|+(n_{t}+p-1)\log |\Sigma_{t}|-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1)),
 \end{split}
 \end{equation}
-where the first two terms depend on $S$, and therefore we need to
-compute the corresponding integral, while for the rest, the integral
-is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
-In order to compute:
+where the first two terms depend on $S$, and therefore we need to compute the corresponding integral, while for the rest, the integral is proportional to the PDF of the matrix--$t$ distribution, Eq.~\eqref{eq:matrixtpdf}.
+
+To compute
 \begin{equation}
-n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
+n_{t}\int_{S\succ 0}j(T_{i},S;\left\{M,\Sigma,\Omega,n\right\})\log |S|\,dS\label{eq:firstlogpart}
 \end{equation}
-we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
-we have:
+we note that taking the derivative with respect to $x$ in Eq.~\eqref{eq:intmultgammadef} we have
 \begin{equation}
-\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
+\int_{S\succ 0}e^{-\textrm{tr}\left[SA\right]}|S|^{x}\log |S|\,dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}}(\log |A|+\gamma_{p}\left(\frac{1+p+2x}{2}\right)).\label{eq:intderivative}
 \end{equation}
-We can solve Eq.~\ref{eq:firstlogpart} by substituting this result directly.
+Equation~\eqref{eq:firstlogpart} can be computed by utilizing this result directly.
 
@@
 \begin{equation}
-\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))dS
+\int_{S\succ 0}j\left(T_{i},S,\left\{M,\Sigma,\Omega,n\right\}\right)\textrm{tr}\left[S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,dS,
 \end{equation}
-we note that, due to the linearity of the trace, the result is equivalent
-to solving a matrix valued integral
+we note that due to the linearity of the trace the result is equivalent to solving a matrix valued integral
 \begin{equation}
-\textrm{tr}\left[\left(\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)SdS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
+\textrm{tr}\left[\left(\int_{S\succ 0}j\left(T_{i},S,\left\{M,\Sigma,\Omega,n\right\}\right)S\,dS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
 \end{equation}
-The enclosed integral also follows from Eq.~\ref{eq:intmultgammadef},
-after taking the matrix derivative with respect to $A$. Using, for
-symmetric $S$
+The enclosed integral also follows from Eq.~\eqref{eq:intmultgammadef}, after taking the matrix derivative with respect to $A$. Using, for symmetric $S$
 \begin{equation}
@@
 \begin{equation}
-\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}SdS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
+\int_{S\succ 0}e^{-\textrm{tr}\left[SA\right]}|S|^{x}S\,dS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
 \end{equation}
@@
 
-Combining results of the two integrals together, and with adequate
-rearrangements we obtain the integrated form of Eq.~\ref{eq:expdef}:
+Combining results of the two integrals together with adequate rearrangements we obtain the integrated form of Eq.~\eqref{eq:expdef}
 \begin{equation}
-\begin{split}E=\sum_{i}^{N}n_{t}\psi_{p}\left(\frac{1}{2}(m+n+p-1)\right)-2\log\Gamma_{p}\left(\frac{1}{2}(n_{t}+p-1)\right)+p\log\det\Omega_{t}^{-1}\\
--n_{t}\log\det\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)+(n_{t}+p-1)\log\det\Sigma_{t}\\
+\begin{split}E=\sum_{i}^{N}n_{t}\psi_{p}\left(\frac{1}{2}(m+n+p-1)\right)-2\log\Gamma_{p}\left(\frac{1}{2}(n_{t}+p-1)\right)+p\log|\Omega_{t}|^{-1}\\
+-n_{t}\log|(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma|+(n_{t}+p-1)\log |\Sigma_{t}|\\
 -(m+n+p-1)\textrm{tr}\left(\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left((T_{i}-M_{t})\Omega_{t}^{-1}(T_{i}-M_{t})^{\textrm{t}}+\Sigma_{t}\right)\right)
@@
 \end{equation}
-The maximization formulas are obtained setting the derivatives of
-this equation to zero.
+The maximization formulas are obtained setting the derivatives of this equation to zero.
 
@@
 \begin{equation}
--p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log\det W_{i}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
+-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\det\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
 \end{equation}

### 2017-01-05 paper.tex
Working on introduction paragraph.

@@
 \begin{abstract}
-We provide expectation-maximization (EM) formulae for the maximum
-likelihood estimation of matrix--$t$ distributions and discuss their
-practical implementation.  The result can be applied to problems where
-the parameters of a matrix--$t$ distribution need to be obtained from
-data, such as the design of recommendation systems, or remote sensing
-algorithms.
+We provide expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
+The result can be applied to problems where the parameters of a matrix--$t$ distribution need to be obtained from data, such as the design of recommendation systems and remote sensing algorithms.
 \end{abstract}
@@
 \paragraph{Introduction}
-Matrix--$t$ distributions have been applied to several predictive
-problems, such as spatial interpolation applied to the prediction of
-pollution concentration \citep{KIBRIA2006785}, recommendation systems
-\citep{NIPS2007_3203}. In these situations it is necessary to obtain
-the parameters of the metrix--$t$ distribution (consisting of the
-number of degrees of freedom, a location matrix and two scale
-matrices) that describe a set of given matrix valued observations.
-While the exact formulae for Maximum Likelihood Estimation (MLE) are
-unwieldy and slow to compute, the problem can be simplified if instead
+Matrix--$t$ distribution is a generalization of multivariate $t$ distribution to matrix variates situations.
+Similar as other $t$ distributions, matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, matrix normal distributions \citep{}, except the mixture parameter is a matrix rather than a scalar.
+Due to this, the distribution shares the importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785}, recommendation systems \citep{NIPS2007_3203}, and ...
+In these problems, it is necessary to obtain the parameters of the distribution, which consist of a degrees of freedom parameter, a matrix location parameter, and two scale matrix parameters, that describe a set of given matrix--valued observations for further inference.
+
+A common approach on estimating the parameters is through the method of maximum likelihood (ML).
+However, optimizing the likelihood directly is a formidable task and the gradient equations are unwieldy and slow to compute and solve.
+For $t$-distributions, it is well-known that one can exploit the scale mixture property and ......
+he problem can be simplified if instead
 we obtain the distribution parameters using Expectation-Maximization

### 2017-01-06 paper.tex
Finished the intro section.

@@
 
-\title{ML Estimation of Matrix t Distributions with Expectation Maximization}
+\title{ML Estimation of Matrix--$t$ Distributions with Expectation Maximization}
 
@@
 \begin{abstract}
-We provide expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
+We provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
 The result can be applied to problems where the parameters of a matrix--$t$ distribution need to be obtained from data, such as the design of recommendation systems and remote sensing algorithms.
@@
 
-\paragraph{Introduction}
+\section{Introduction}
 Matrix--$t$ distribution is a generalization of multivariate $t$ distribution to matrix variates situations.
@@
 A common approach on estimating the parameters is through the method of maximum likelihood (ML).
-However, optimizing the likelihood directly is a formidable task and the gradient equations are unwieldy and slow to compute and solve.
-For $t$-distributions, it is well-known that one can exploit the scale mixture property and ......
-he problem can be simplified if instead
-we obtain the distribution parameters using Expectation-Maximization
-(EM). We obtain can obtain the EM expressions by viewing the
-matrix--$t$  distribution as a matrix-normal distribution where one of
-the scale matrices is sampled from a Wishart distribution.
+However, optimizing the likelihood directly is a formidable task as the gradient equations are unwieldy slow to compute and solve.
+For $t$ distributions, it is well-known that one can exploit the property of scale mixture and apply expectation maximization (EM) technique to optimize the likelihood function \citep{}.
+The same trick can be also applied to matrix--$t$ distribution by viewing it a matrix-normal distribution with inverse Wishart distributed scale parameter.
 
-\paragraph{Derivation}
+This note is organized as follows.
+In section \ref{sec:derivation}, we present an EM formulae for MLE of matrix--$t$ distribution together with the relevant details on the derivation.
+The notation and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section.
+Practical implementation of the algorithm is discussed in section \ref{sec:implement} and a brief conclusion is provided in section \ref{sec:conclusion}.
+
+\section{EM formulae for matrix--$t$ distribution}\label{sec:derivation}
 
@@
 with Wishart distributed row scale parameter $S \sim W_{p}\left(\Sigma,\nu\right)$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale
-parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.
+parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.
 
@@
 \end{equation}
-The maximization formulas are obtained setting the derivatives of this equation to zero.
+The maximization formulae is obtained setting the derivatives of this equation to zero.
 
@@
 
+\section{Implementation}\label{sec:implement}
+
+\section{Conclusion}\label{sec:conclusion}
+
 \bibliographystyle{unsrtnat}

### 2017-01-09 paper.tex
Added Acknowledgement
Starting to organize section 2 to make it more readable.

@@
 Matrix--$t$ distribution is a generalization of multivariate $t$ distribution to matrix variates situations.
-Similar as other $t$ distributions, matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, matrix normal distributions \citep{}, except the mixture parameter is a matrix rather than a scalar.
+Similar as other $t$ distributions, matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, matrix normal distributions, except the mixture parameter is a matrix rather than a scalar \citep{gupta1999matrix}.
 Due to this, the distribution shares the importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785}, recommendation systems \citep{NIPS2007_3203}, and ...
@@
 However, optimizing the likelihood directly is a formidable task as the gradient equations are unwieldy slow to compute and solve.
-For $t$ distributions, it is well-known that one can exploit the property of scale mixture and apply expectation maximization (EM) technique to optimize the likelihood function \citep{}.
-The same trick can be also applied to matrix--$t$ distribution by viewing it a matrix-normal distribution with inverse Wishart distributed scale parameter.
+For $t$ distributions, it is well-known that one can exploit the property of scale mixture and apply expectation maximization (EM) technique to optimize the likelihood function \citep{10.2307/24305551}.
+The same trick can be also applied to matrix--$t$ distribution by viewing it as a matrix-normal distribution with inverse Wishart distributed scale parameter.
 
@@
 The notation and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section.
-Practical implementation of the algorithm is discussed in section \ref{sec:implement} and a brief conclusion is provided in section \ref{sec:conclusion}.
+Practical implementation of the algorithm is discussed in section \ref{sec:implement} and a brief conclusion is given in section \ref{sec:conclusion}.
 
@@
 
-The derivation of the EM formulae is mainly based on the following
-integral identity:
+All matrices referred in this note are real.
+For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\left\{S_p\succ 0\right\}$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
 
+We first review some basic definitions on matrix--$t$ distributions.
+A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $n>0$.
+A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,n)$ with location parameter $M$ if given the row scale parameter $S$, $T$ is matrix normal distributed
+
+The derivation of the EM formulae is mainly based on the integral identity \citep{gupta1999matrix}
 \begin{equation}
-\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}\,dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
+\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}\,dS=\Gamma_{p}\left(x+\frac{1+p}{2}\right)|A|^{x+\frac{1+p}{2}}.\label{eq:intmultgammadef}
 \end{equation}
-where the domain of integration is the set of real symmetric
+Here the domain of integration is the set of real symmetric
 positive definite matrices with dimension $p$, $A$ is a $p\times p$
@@
 \end{equation}
-For notation convenience, the matrix determinant is represented as
-$|\cdot|$. We also define the multivariate digamma function
+We also define the multivariate digamma function
 %
@@
 
+\section*{Acknowledgement}
+This work is not supported by any grant and has not been presented in any regional or international meetings.
+
 \bibliographystyle{unsrtnat}

### 2017-01-12 paper.tex
Working on section 2.

@@
 For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\left\{S_p\succ 0\right\}$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
-
-We first review some basic definitions on matrix--$t$ distributions.
-A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $n>0$.
-A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,n)$ with location parameter $M$ if given the row scale parameter $S$, $T$ is matrix normal distributed
-
-The derivation of the EM formulae is mainly based on the integral identity \citep{gupta1999matrix}
+The multivariate gamma
+function, $\Gamma_{p}(\cdot)$, given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
 \begin{equation}
-\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}\,dS=\Gamma_{p}\left(x+\frac{1+p}{2}\right)|A|^{x+\frac{1+p}{2}}.\label{eq:intmultgammadef}
-\end{equation}
-Here the domain of integration is the set of real symmetric
-positive definite matrices with dimension $p$, $A$ is a $p\times p$
-nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma
-function, given in terms of the standard gamma function $\Gamma(\cdot)$
-by
-\begin{equation}
-\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
+\Gamma_{p}(x)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(x+\frac{1-j}{2}\right).\label{eq:multgammadef}
 \end{equation}
@@
 %
-\begin{equation} \psi_{p}(a)=\frac{\partial}{\partial
-	a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
+\begin{equation}
+\psi_{p}(x):=\frac{\partial}{\partial x}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
 \end{equation}
 
-A random $p\times m$ matrix $T$ following a matrix--$t$ distribution
-$T_{p,m}(M,\Sigma,\Omega,n)$ has the probability density
+We first review some basic definitions on matrix--$t$ distributions.
+A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $\nu>0$.
+A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,\nu)$ with location parameter $M$ if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the random matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
 \begin{equation}
-f(T;\left\{M,\Sigma,\Omega,n\right\})=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
-|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-n}{2}},
+T|M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
+\end{equation}
+and
+\begin{equation}
+S|\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
+\end{equation}
+
+The matrix normal distribution $N_{p,m}(M,S,\Omega)$ has density
+\begin{equation}
+\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
+\end{equation}
+at $X$ while the inverse Wishart distribution $W^{-1}(\Sigma,n)$ with $n>p-1$ has density
+\begin{equation}
+\frac{|\Sigma|^{\frac{n}{2}}}{2^{\frac{n p}{2}}|S|^{\frac{n+p+1}{2}}\Gamma_{p}(\frac{n}{2})}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Sigma S^{-1}\right]\right)\label{eq:wdistpdf}
+\end{equation}
+at $S$. Here $\Gamma_{p}(\cdot)$ is the
+multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.
+
+Follow the definition in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
+\begin{equation}
+\frac{\Gamma_{p}\left(\frac{p+\nu+m-1}{2}\right)}{\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
+|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-\nu}{2}}
 \label{eq:matrixtpdf}
 \end{equation}
-where $\Sigma$ and $\Omega$ are both real symmetric positive definite
-matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$
-matrix that indicates the location of the distribution, and $n$ is the
-number of degrees of freedom.
+with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant
+\begin{equation}
+...
+\end{equation}
+and the integral identity \citep{gupta1999matrix}
+\begin{equation}
+\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}},\label{eq:intmultgammadef}
+\end{equation}
+where $A$ is a $p\times p$ nonsingular matrix.
 
-Following \citet{gupta1999matrix}, we begin by deriving the
-probability density of the matrix--$t$ distribution starting from
-a matrix normal distributed random variate $X\sim N_{p,m}(M,S,\Omega)$
-with Wishart distributed row scale parameter $S \sim W_{p}\left(\Sigma,\nu\right)$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale
-parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.
-
-The $p\times m$ random matrix $X \sim N_{p,m}(M,S,\Omega)$ has density
-\begin{equation}
-	\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right),\label{eq:ndistpdf}
-\end{equation}
-where $S$ and $\Omega$ are symmetric positive definite matrices of dimension $p$ and $m$, respectively, and $M$ (the expectation of $X$) is a $p\times m$ matrix.
-The matrix $S$ is proportional to the covariance of the rows of $X$ while the matrix $\Omega$ is proportional to the covariance of the columns of $X$.
-In turn the $p\times p$ matrix $S\sim W_{p}\left(\Sigma,\nu\right)$ has density
-\begin{equation}
-\frac{|S|^{\frac{\nu-p-1}{2}}}{2^{\nu p/2}|\Sigma|^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}\exp\left(-\textrm{tr}\left[\Sigma^{-1}S\right]\right),\label{eq:wdistpdf}
-\end{equation}
-where $\Sigma$ is a $p\times p$ symmetric positive definite matrix, $\nu>p-1$
-is the number of degrees of freedom, and $\Gamma_{p}$ is the
-multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.
+We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.
 

### 2017-01-12 paper.tex
Finished editing section 2.1.

@@
 
-We first review some basic definitions on matrix--$t$ distributions.
+\subsection{Matrix--$t$ distribution}
+
 A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $\nu>0$.
-A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed $T_{p,m}(M,\Sigma,\Omega,\nu)$ with location parameter $M$ if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the random matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
+A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
 \begin{equation}
@@
 \end{equation}
-
 The matrix normal distribution $N_{p,m}(M,S,\Omega)$ has density
@@
 \end{equation}
-with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant
+with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant for inverse SPD matrix $U=S^{-1}$
 \begin{equation}
-...
+\left|\frac{\partial U}{\partial S}\right|=|S|^{-p-1}
 \end{equation}
@@
 \begin{equation}
-\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}},\label{eq:intmultgammadef}
+\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}},\label{eq:intmultgammadef}
 \end{equation}
@@
 
-We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulae.
-
-We will show that the matrix
-\begin{equation}
-T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
-\end{equation}
-follows a matrix--$t$ distribution with degrees of freedom $n=\nu-p+1$.
-Here $S^{\frac{1}{2}}$ is the lower Cholesky factor of $S$, i.e., $S=S^{\frac{1}{2}}(S^{\frac{1}{2}})^t$.
-The Jacobian determinant of the transformation \eqref{eq:change} is
-\begin{equation}
-\left|\frac{\partial T}{\partial X}\right|=|S|^{-\frac{m}{2}}.\label{eq:jac}
-\end{equation}
-Using Eq.~\eqref{eq:ndistpdf}, Eq.~\eqref{eq:wdistpdf} and Eq.~\eqref{eq:jac} we can see the joint density of $T$ and $S$ is given as
-\begin{equation}
-\begin{split}j(T,S;\left\{M,\Sigma,\Omega,n\right\})=\frac{\pi^{-\frac{mp}{2}}|\Sigma|^{\frac{n+p-1}{2}}|\Omega|^{\frac{n+m-2}{2}}|S|^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
-\exp\left(-\frac{1}{2}\textrm{tr}\left[S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right]\right).
-\end{split}
-\label{eq:jointpdf}
-\end{equation}
-
-Finally we obtain the density \eqref{eq:matrixtpdf} of the matrix--$t$ distribution
-by integrating over $S$ in the above expression, using Eq.~\eqref{eq:change} and slightly
-rearranging the terms. In particular noting that
-$|\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|=|\Sigma||I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|$.
+Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant, i.e., $T_{p,m}(M,c\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any $c>0$.
 

### 2017-01-14 paper.tex
Working on the subsection for EM.

@@
 \usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage{amsthm}
@@
 
-\section{EM formulae for matrix--$t$ distribution}\label{sec:derivation}
+\section{EM for matrix--$t$ distribution}\label{sec:derivation}
 
@@
 
-\subsection{Calculating the Expectation-maximization steps}
+\subsection{EM formulas for matrix--$t$ distribution}
+
+Given a random variable $X$ with probability distribution with density (or mass) function $p_{\theta}(\cdot)$ with $\theta$ dependence.
+Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
+With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method of MLE consists of two step:
+
+\begin{itemize}
+\item At the $n$-th iteration, compute the conditional expectations with respect to $Z$
+\begin{equation}
+Q(\theta|\theta_n):=\sum_{i\in I}\mathbb{E}_{\theta_n}\left[\log p_{\theta}(x_i,Z)|x_i\right].
+\end{equation}
+
+\item Update $\theta$ by maximizing $Q(\theta|\theta_n)$, i.e., $\theta_{n+1} = \textrm{argmax}_{\theta}\,Q(\theta|\theta_n)$.
+\end{itemize}
+
+For matrix--$t$ distributions,
 

### 2017-01-14 paper.tex
Still Working on EM subsection.

@@
 
-Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant, i.e., $T_{p,m}(M,c\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any $c>0$.
+Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant, i.e., $T_{p,m}(M,c\,\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any constant $c>0$.
 
@@
 Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
-With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method of MLE consists of two step:
+With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method of MLE consists of two steps:
 
 \begin{itemize}
-\item At the $n$-th iteration, compute the conditional expectations with respect to $Z$
+\item At the $k$-th iteration, compute the conditional expectations with respect to $Z$
 \begin{equation}
-Q(\theta|\theta_n):=\sum_{i\in I}\mathbb{E}_{\theta_n}\left[\log p_{\theta}(x_i,Z)|x_i\right].
+Q(\theta,\theta_k):=\sum_{i\in I}\mathbb{E}_{\theta_k}\left[\log p_{\theta}(x_i,Z)|x_i\right].
 \end{equation}
 
-\item Update $\theta$ by maximizing $Q(\theta|\theta_n)$, i.e., $\theta_{n+1} = \textrm{argmax}_{\theta}\,Q(\theta|\theta_n)$.
+\item Update $\theta$ by maximizing $Q(\theta,\theta_k)$, i.e., $\theta_{k+1} = \textrm{argmax}_{\theta}\,Q(\theta,\theta_k)$.
 \end{itemize}
 
-For matrix--$t$ distributions,
+Given observed matrix data $\left\{T_j\right\}_{j\in J}$, for each observation $T_j$ the E-step of matrix--$t$ distributions entails computing the following integral at step $k$
 
-The expectation maximization formalism entails calculating 
 \begin{equation}
-E=\sum_{i=1}^{N}\frac{\int_{S\succ 0}j(T_{i},S,M,\Sigma,\Omega,n)\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\,dS}{f(T;\left\{M,\Sigma,\Omega,n\right\})},\label{eq:expdef}
+Q(\theta,\theta_k)=\frac{\int_{S\succ 0}f(T_{j},S,\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\})\log f\left(T_{j},S,\left\{M,\Sigma,\Omega,\nu\right\}\right)\,dS}{f(T_j;\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\})},\label{eq:expdef}
 \end{equation}
-where $\{T_i\}_{i=1}^N$ is a dataset consists of matrix elements $T_i$ from which we wish to estimate the maximum likelihood parameters of the distribution.
-We omit writing explicitly the dependence of $E$ on all the parameters.
-The values of the new guesses are found by maximizing with respect to the parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$.
 

### 2017-01-15 paper.tex
Modifying the derivations as I used a slighlty different definition to define matrix-t.

@@
 All matrices referred in this note are real.
-For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\left\{S_p\succ 0\right\}$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
+For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_+^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
 The multivariate gamma
@@
 \begin{equation}
-T|M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
+T\,|\,M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
 \end{equation}
@@
 \begin{equation}
-S|\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
+S\,|\,\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
 \end{equation}
@@
 \begin{equation}
-\int_{\left\{S_p\succ 0\right\}}e^{-\textrm{tr}[S_pA]}|S_p|^{\alpha}\,dS_p=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}},\label{eq:intmultgammadef}
+\int_{\mathbf{S}_+^p}e^{-\textrm{tr}[S A]}|S|^{\alpha}\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}},\label{eq:intmultgammadef}
 \end{equation}
@@
 Given observed matrix data $\left\{T_j\right\}_{j\in J}$, for each observation $T_j$ the E-step of matrix--$t$ distributions entails computing the following integral at step $k$
+\begin{equation}
+Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_+^p}f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS,\label{eq:expdef}
+\end{equation}
+where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$, $\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and $f(T,S;\theta)$ is the full likelihood of the matrix--$t$ distribution with the hidden scale parameter $S$.
 
+We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log f(T_{j},S;\theta)$.
+Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the parameters in $\theta$ we have
 \begin{equation}
-Q(\theta,\theta_k)=\frac{\int_{S\succ 0}f(T_{j},S,\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\})\log f\left(T_{j},S,\left\{M,\Sigma,\Omega,\nu\right\}\right)\,dS}{f(T_j;\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\})},\label{eq:expdef}
+\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}(\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}})\right]\\
+-\nu p\log2-p\log |\Omega|+(\nu+p-1)\log |\Sigma|-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right).
+\end{split}\label{eq:logf}
 \end{equation}
-
-We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the
-maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$ we have
+From Eq.~\eqref{eq:logf}, we can see the only integrals which need to be computed are
 \begin{equation}
-\begin{split}\log j(T_{i},S;\left\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\right\})\sim n_{t}\log|S|+\textrm{tr}\left[S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\\
--n_{t}p\log2-p\log |\Omega_{t}|+(n_{t}+p-1)\log |\Sigma_{t}|-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1)),
-\end{split}
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS
 \end{equation}
-where the first two terms depend on $S$, and therefore we need to compute the corresponding integral, while for the rest, the integral is proportional to the PDF of the matrix--$t$ distribution, Eq.~\eqref{eq:matrixtpdf}.
+and
+\begin{equation}
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.
+\end{equation}
 

### 2017-01-15 paper.tex
Working on the em section.

@@
 \begin{equation}
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS\label{eq:firstlogpart}
 \end{equation}
@@
 \begin{equation}
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.\label{eq:2ndlogpart}
 \end{equation}
-
-To compute
+We note that taking the derivative with respect to $\alpha$ in Eq.~\eqref{eq:intmultgammadef} we have
 \begin{equation}
-n_{t}\int_{S\succ 0}j(T_{i},S;\left\{M,\Sigma,\Omega,n\right\})\log |S|\,dS\label{eq:firstlogpart}
+\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}}\left\{\log |A|+\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)\right\}\label{eq:intderivative}
 \end{equation}
-we note that taking the derivative with respect to $x$ in Eq.~\eqref{eq:intmultgammadef} we have
-\begin{equation}
-\int_{S\succ 0}e^{-\textrm{tr}\left[SA\right]}|S|^{x}\log |S|\,dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}}(\log |A|+\gamma_{p}\left(\frac{1+p+2x}{2}\right)).\label{eq:intderivative}
-\end{equation}
-Equation~\eqref{eq:firstlogpart} can be computed by utilizing this result directly.
-
-In order to compute the second nontrivial log term,
+and the integral \eqref{eq:firstlogpart} can be computed by utilizing this result directly. To compute the second nontrivial log term \eqref{eq:2ndlogpart}
 \begin{equation}

### 2017-01-20 paper.tex
Adjusted the E-step expressions.

@@
 \end{equation}
-We note that taking the derivative with respect to $\alpha$ in Eq.~\eqref{eq:intmultgammadef} we have
+Note that taking the derivative with respect to $\alpha$ in Eq.~\eqref{eq:intmultgammadef} we have
 \begin{equation}
-\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}}\left\{\log |A|+\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)\right\}\label{eq:intderivative}
+\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}}\left\{\log |A|+\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)\right\}\label{eq:intderivative1}
 \end{equation}
-and the integral \eqref{eq:firstlogpart} can be computed by utilizing this result directly. To compute the second nontrivial log term \eqref{eq:2ndlogpart}
+and hence the integral \eqref{eq:firstlogpart} follows this result directly.
+To compute the integral \eqref{eq:2ndlogpart} we take the matrix derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $A$ and obtain
 \begin{equation}
-\int_{S\succ 0}j\left(T_{i},S,\left\{M,\Sigma,\Omega,n\right\}\right)\textrm{tr}\left[S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,dS,
-\end{equation}
-we note that due to the linearity of the trace the result is equivalent to solving a matrix valued integral
-\begin{equation}
-\textrm{tr}\left[\left(\int_{S\succ 0}j\left(T_{i},S,\left\{M,\Sigma,\Omega,n\right\}\right)S\,dS\right)(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}})\right]\,.
-\end{equation}
-The enclosed integral also follows from Eq.~\eqref{eq:intmultgammadef}, after taking the matrix derivative with respect to $A$. Using, for symmetric $S$
-\begin{equation}
-\frac{d}{dA}\det A=(\det A)A^{-1}\label{eq:detderivative}
-\end{equation}
-and
-\begin{equation}
-\frac{d}{dA}e^{\textrm{tr}SA}=(e^{\textrm{trS}A})S
-\end{equation}
-we have 
-\begin{equation}
-\int_{S\succ 0}e^{-\textrm{tr}\left[SA\right]}|S|^{x}S\,dS=-\frac{1}{2}(1+p+2x)\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(p+2x-1)}A^{-1}
+\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=-\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{p-1}{2}}A^{-1},\label{eq:intderivative2}
 \end{equation}
@@
 
-Combining results of the two integrals together with adequate rearrangements we obtain the integrated form of Eq.~\eqref{eq:expdef}
+Combining results of Eqs.~\eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form of $Q_j(\theta,\theta_k)$ while discarding the irrelevant constant terms
 \begin{equation}
-\begin{split}E=\sum_{i}^{N}n_{t}\psi_{p}\left(\frac{1}{2}(m+n+p-1)\right)-2\log\Gamma_{p}\left(\frac{1}{2}(n_{t}+p-1)\right)+p\log|\Omega_{t}|^{-1}\\
--n_{t}\log|(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma|+(n_{t}+p-1)\log |\Sigma_{t}|\\
--(m+n+p-1)\textrm{tr}\left(\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left((T_{i}-M_{t})\Omega_{t}^{-1}(T_{i}-M_{t})^{\textrm{t}}+\Sigma_{t}\right)\right)
+\begin{split}Q_j(\theta,\theta_k)\sim \nu\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)+p\log|\Omega|^{-1}\\
+-\nu\log|(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}+\Sigma_k|+(\nu+p-1)\log |\Sigma|\\
+-(m+\nu_k+p-1)\textrm{tr}\left[\left((T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}+\Sigma_k\right)^{-1}\right.\\
+\left.\left((T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}+\Sigma\right)\right].
 \end{split}

### 2017-01-21 paper.tex
Finished checking the E-step and corrected some minor typos.

@@
 \begin{equation}
-Q(\theta,\theta_k):=\sum_{i\in I}\mathbb{E}_{\theta_k}\left[\log p_{\theta}(x_i,Z)|x_i\right].
+Q(\theta,\theta_k)=\sum_{i\in I}Q_i(\theta,\theta_k),\qquad Q_i(\theta,\theta_k)=\mathbb{E}_{\theta_k}\left[\log p_{\theta}(x_i,Z)|x_i\right].
 \end{equation}
@@
 \begin{equation}
-\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}(\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}})\right]\\
+\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]\\
 -\nu p\log2-p\log |\Omega|+(\nu+p-1)\log |\Sigma|-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right).
@@
 \end{equation}
-From Eq.~\eqref{eq:logf}, we can see the only integrals which need to be computed are
+The integrals to be computed are
 \begin{equation}
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS\label{eq:firstlogpart}
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS
 \end{equation}
-and
+and they can be done by using the equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively,
 \begin{equation}
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.\label{eq:2ndlogpart}
+\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
 \end{equation}
-Note that taking the derivative with respect to $\alpha$ in Eq.~\eqref{eq:intmultgammadef} we have
 \begin{equation}
-\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{1+p}{2}}\left\{\log |A|+\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)\right\}\label{eq:intderivative1}
+\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
 \end{equation}
-and hence the integral \eqref{eq:firstlogpart} follows this result directly.
-To compute the integral \eqref{eq:2ndlogpart} we take the matrix derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $A$ and obtain
+
+Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
 \begin{equation}
-\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=-\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{\alpha+\frac{p-1}{2}}A^{-1},\label{eq:intderivative2}
-\end{equation}
-from where again, we can obtain the result by direct substitution.
-
-Combining results of Eqs.~\eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form of $Q_j(\theta,\theta_k)$ while discarding the irrelevant constant terms
-\begin{equation}
-\begin{split}Q_j(\theta,\theta_k)\sim \nu\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)+p\log|\Omega|^{-1}\\
--\nu\log|(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}+\Sigma_k|+(\nu+p-1)\log |\Sigma|\\
--(m+\nu_k+p-1)\textrm{tr}\left[\left((T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}+\Sigma_k\right)^{-1}\right.\\
-\left.\left((T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}+\Sigma\right)\right].
+\begin{split}Q_j(\theta,\theta_k) \sim \nu\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
+-\nu\log|\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}|+(\nu+p-1)\log |\Sigma|\\
+-(m+\nu_k+p-1)\textrm{tr}\left[\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}\right.\\
+\left.\left\{(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}+\Sigma\right\}\right].
 \end{split}
@@
 \end{equation}
-The maximization formulae is obtained setting the derivatives of this equation to zero.
+The maximization formulae follows by setting the gradient of the parameters to zero.
 

### 2017-01-21 paper.tex
Start modifying the M-step.

@@
 Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
-With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method of MLE consists of two steps:
+With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method for MLE consists of two steps:
 
@@
 \int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS
+\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.
 \end{equation}
-and they can be done by using the equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively,
+They can be done by using the equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively,
 \begin{equation}
@@
 \begin{equation}
-\begin{split}Q_j(\theta,\theta_k) \sim \nu\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
--\nu\log|\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}|+(\nu+p-1)\log |\Sigma|\\
--(m+\nu_k+p-1)\textrm{tr}\left[\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}\right.\\
-\left.\left\{(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}+\Sigma\right\}\right].
+\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
+-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
 \end{split}
@@
 \end{equation}
-The maximization formulae follows by setting the gradient of the parameters to zero.
-
-For $M_{t}$ we have:
+where the matrices $W_j^{(k)}$ are defined as
 \begin{equation}
-\frac{dE}{dM_{t}}=0\Rightarrow\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left(T_{i}-M_{t}\right)\Omega_{t}^{-1}=0
-\end{equation}
-and because $\Omega_{t}$ is nonsingular, we can obtain $M_{t}$ as
-a weighted average of the input data:
-\begin{equation}
-M_{t}=\frac{1}{N}\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}T_{i}=\frac{1}{N}\sum_{i}^{N}W_{i}T_{i}
-\end{equation}
-where we have defined the weight matrices $W_{i}$ as
-\begin{equation}
-W_{i}=\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}
+W_j^{(k)}=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
 \end{equation}
 
-$\Sigma_{t}$ is given in terms of the inverse of the mean weight
-by:
+To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
 \begin{equation}
-\Sigma_{t}=\frac{n_{t}+p-1}{n+m+p-1}\left[\frac{1}{N}\sum_{i}^{N}W_{i}\right]^{-1}
+\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
 \end{equation}
-Note that it depends on $n_{t}$.
+As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
+\begin{equation}
+M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.
+\end{equation}
+Maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{k}}$
+\begin{equation}
+\Sigma_{k+1}=\frac{\nu+p-1}{\nu_k+m+p-1}\left(\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\right)^{-1}. 
+\end{equation}
+Note that $\Sigma_{k+1}$ given above depends on $\nu$.
 
-For $\Omega_{t}$ we obtain:
+For $\Omega_{k+1}$ we obtain
 \begin{equation}

### 2017-01-21 paper.tex
Corrected some typos in M-step simplified formulas.

@@
 \begin{equation}
-\Sigma_{k+1}=\frac{\nu+p-1}{\nu_k+m+p-1}\left(\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\right)^{-1}. 
+\Sigma_{k+1}=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\right)^{-1}. 
 \end{equation}
-Note that $\Sigma_{k+1}$ given above depends on $\nu$.
+Note that $\Sigma_{k+1}$ given above has dependence on $\nu$. Similarly for $\Omega$ we obtain
+\begin{equation}
+\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).
+\end{equation}
 
-For $\Omega_{k+1}$ we obtain
+For $\nu$ in the case when $\Sigma$ is a free parameter to be estimated we have
 \begin{equation}
-\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
-\end{equation}
-and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
-that needs to be estimated, we have:
-\begin{equation}
--p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\det\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
+p\log\left(\frac{n_{t}+p-1}{m+n+p-1}\right)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}\left(\frac{m+n+p-1}{2}\right)-\psi_{p}\left(\frac{n_{t}+p-1}{2}\right)=0
 \end{equation}
@@
 \begin{equation}
-\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
+\log|\Sigma_{t}|-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
 \end{equation}

### 2017-01-22 paper.tex
Finished editing the EM section.

@@
 
-\subsection{EM formulas for matrix--$t$ distribution}
+\subsection{EM formulae for matrix--$t$ distribution}
 
@@
 \begin{equation}
-W_j^{(k)}=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
+W_j^{(k)}:=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
 \end{equation}
@@
 \end{equation}
-Maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{k}}$
-\begin{equation}
-\Sigma_{k+1}=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\right)^{-1}. 
-\end{equation}
-Note that $\Sigma_{k+1}$ given above has dependence on $\nu$. Similarly for $\Omega$ we obtain
+Maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$ we obtain $\Omega_{k+1}$
 \begin{equation}
@@
 \end{equation}
-
-For $\nu$ in the case when $\Sigma$ is a free parameter to be estimated we have
+For the rest of the parameter(s), we consider two possible situations.
+\begin{itemize}
+\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{(k)}}$
 \begin{equation}
-p\log\left(\frac{n_{t}+p-1}{m+n+p-1}\right)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}\left(\frac{m+n+p-1}{2}\right)-\psi_{p}\left(\frac{n_{t}+p-1}{2}\right)=0
+\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:sigma}
 \end{equation}
-otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
-$n_{t}$:
+$\Sigma_{k+1}$ has linear dependence on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
+\[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
+is a concave function of $\nu$ for all $\nu>0$.
+The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive.
+For this case we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$, which has the form
 \begin{equation}
-\log|\Sigma_{t}|-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
+h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0
 \end{equation}
+with $h(x):=p\log(x)-\psi_p(x/2)$.
+\item $\Sigma$ is known and $\nu$ is unknown: In this situation, $Q\left(\left\{M_{k+1},\Sigma,\Omega_{k+1},\nu\right\},\theta_k\right)$ is also a concave function of $\nu$.
+The gradient equation of $\nu$ is given by
+\begin{equation}
+\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.
+\end{equation}
+\end{itemize}
 
 \section{Implementation}\label{sec:implement}
+In the actual implementation of
 

### 2017-01-22 paper.tex
Arranged the sections.

@@
 This note is organized as follows.
-In section \ref{sec:derivation}, we present an EM formulae for MLE of matrix--$t$ distribution together with the relevant details on the derivation.
-The notation and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section.
-Practical implementation of the algorithm is discussed in section \ref{sec:implement} and a brief conclusion is given in section \ref{sec:conclusion}.
+In section \ref{sec:matrixt} we give a brief review on matrix--$t$ distributions.
+The notations and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section.
+EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief conclusion is given in section \ref{sec:conclusion}.
 
-\section{EM for matrix--$t$ distribution}\label{sec:derivation}
+\section{Matrix--$t$ distributions}\label{sec:matrixt}
 
@@
 For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_+^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
-The multivariate gamma
-function, $\Gamma_{p}(\cdot)$, given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
+The multivariate gamma function $\Gamma_{p}(\cdot)$ given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
 \begin{equation}
@@
 \begin{equation}
-\psi_{p}(x):=\frac{\partial}{\partial x}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
+\psi_{p}(x):=\frac{d}{dx}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
 \end{equation}
-
-\subsection{Matrix--$t$ distribution}
 
@@
 
-\subsection{EM formulae for matrix--$t$ distribution}
+\section{EM formulae for matrix--$t$ distribution}\label{sec:EM}
 
@@
 \end{itemize}
+
+\subsection{Expectation step}
 
@@
 \end{equation}
-where the matrices $W_j^{(k)}$ are defined as
+where the matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ are defined as
 \begin{equation}
@@
 \end{equation}
+
+\subsection{Maximization step}
 
@@
 \begin{equation}
-M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.
+M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.\label{eq:M}
 \end{equation}
@@
 \begin{equation}
-\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).
+\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).\label{eq:Omega}
 \end{equation}
@@
 \begin{equation}
-\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:sigma}
+\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:Sigma}
 \end{equation}
@@
 \begin{equation}
-h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0
+h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0\label{eq:nu1}
 \end{equation}
@@
 \begin{equation}
-\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.
+\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.\label{eq:nu2}
 \end{equation}
 \end{itemize}
-
-\section{Implementation}\label{sec:implement}
-In the actual implementation of
 

### 2017-01-22 paper.tex
Start adding some implementation info.
Mentioned about Sherman-Morrison-Woodbury formula.

@@
 \end{equation}
+As $\Sigma_k$ and $\Omega_k$ are both SPD, each matrix $W_j^{(k)}$ is also SPD.
+The computation of $W_j^{(k)}$ can be done efficiently using Cholesky factorization via LAPACK routines and the BLAS routines for symmetric matrix update.
+When $p$ is much greater than $m$, i.e., the dimension of $\Sigma$ is much larger than $\Omega$, one can apply Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to reduce the cost of inverting the $p\times p$ matrix for each $W_j^{(k)}$.
+Using the formulae $W_j^{(k)}$ can be rewritten in the equivalent form
+\begin{equation}
+W_j^{(k)}=\Sigma_k^{-1}-\Sigma_k^{-1}U_j\left(I_m+U_j^{\textrm{t}}\Sigma_k^{-1}U_j\right)^{-1}U_j^{\textrm{t}}\Sigma_k^{-1}\label{eq:SMW}
+\end{equation}
+with $U_j=(T_j-M_k)\Omega_k^{-1/2}$ and $\Omega_k^{1/2}$ is the upper Cholesky factor of $\Omega_k$.
+Note that in Eq.~\eqref{eq:SMW} one inverts a $m\times m$ matrix for each $j$ and the inversion (or factorization) of $\Sigma_k$ only needs to be done once.
 

### 2017-01-22 paper.tex
Mentioned about fixing the scaling constant.

@@
 The notations and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section.
-EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief conclusion is given in section \ref{sec:conclusion}.
+EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief final discussion is given in section \ref{sec:discussion}.
 
@@
 with $U_j=(T_j-M_k)\Omega_k^{-1/2}$ and $\Omega_k^{1/2}$ is the upper Cholesky factor of $\Omega_k$.
-Note that in Eq.~\eqref{eq:SMW} one inverts a $m\times m$ matrix for each $j$ and the inversion (or factorization) of $\Sigma_k$ only needs to be done once.
+Note that in Eq.~\eqref{eq:SMW} one inverts a $m\times m$ matrix for each $j$ and the inversion (or factorization) of $\Sigma_k$ only needs to be done once in each iteration.
 
@@
 \end{equation}
-$\Sigma_{k+1}$ has linear dependence on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
+$\Sigma_{k+1}$ depends linearly on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
 \[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
@@
 
-\section{Conclusion}\label{sec:conclusion}
+When $\Omega$ and $\Sigma$ are both unknown, the updating steps given in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} make the likelihood of matrix--$t$ distribution increases monotonically but the parameters $\Omega_k$ and $\Sigma_k$ may not converge as $k$ increases due to the non-identifiability, matrix distributions are defined up to a multiplication constant.
+This issue has also been mentioned in \citep{2013arXiv1309.6609G} for the estimation of matrix normal distributions.
+However, instead of fixing a single component in one of the scale matrices, we suggest setting the mean trace ratio
+\begin{equation}
+\frac{\textrm{tr}[\Omega_k]/m}{\textrm{tr}[\Sigma_k]/p}
+\end{equation}
+of $\Omega_k$ and $\Sigma_k$ to 1 with proper scaling at the end of each M-step.
+This ensures the convergence of the algorithm and also avoids possible numerical issues due to the large scale separation between $\Omega_k$ and $\Sigma_k$ in the iterations.
+
+\subsection{Parameters initialization}
+
+\section{Discussion}\label{sec:discussion}
 

### 2017-01-22 paper.tex
Added a short conclusion.

--- 2017-01-22 paper.tex
@@
 \begin{abstract}
-We provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
+We provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss its practical implementation.
 The result can be applied to problems where the parameters of a matrix--$t$ distribution need to be obtained from data, such as the design of recommendation systems and remote sensing algorithms.
@@
 \section{Introduction}
-Matrix--$t$ distribution is a generalization of multivariate $t$ distribution to matrix variates situations.
+The matrix--$t$ distribution is a generalization of the multivariate $t$ distribution to matrix variates situations.
 Similar as other $t$ distributions, matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, matrix normal distributions, except the mixture parameter is a matrix rather than a scalar \citep{gupta1999matrix}.
-Due to this, the distribution shares the importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785}, recommendation systems \citep{NIPS2007_3203}, and ...
+Due to this, the distribution shares its importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785} and recommendation systems \citep{NIPS2007_3203}.
 In these problems, it is necessary to obtain the parameters of the distribution, which consist of a degrees of freedom parameter, a matrix location parameter, and two scale matrix parameters, that describe a set of given matrix--valued observations for further inference.
@@
 For $t$ distributions, it is well-known that one can exploit the property of scale mixture and apply expectation maximization (EM) technique to optimize the likelihood function \citep{10.2307/24305551}.
-The same trick can be also applied to matrix--$t$ distribution by viewing it as a matrix-normal distribution with inverse Wishart distributed scale parameter.
+The same technique can be also applied to matrix--$t$ distribution by viewing it a matrix-normal distribution with inverse Wishart distributed scale parameter.
 
@@
 
-Given a random variable $X$ with probability distribution with density (or mass) function $p_{\theta}(\cdot)$ with $\theta$ dependence.
+Given a random variable $X$ with probability density (or mass) function $p_{\theta}(\cdot)$.
 Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
@@
 \end{equation}
-where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$, $\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and $f(T,S;\theta)$ is the full likelihood of the matrix--$t$ distribution with the hidden scale parameter $S$.
+where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$, $\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and $f(T,S;\theta)$ is the full likelihood of the distribution with the hidden scale parameter $S$.
 
@@
 \end{equation}
-They can be done by using the equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively,
+They can be done by using equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively
 \begin{equation}
@@
 \begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
--(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
+-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|.
 \end{split}
@@
 \end{equation}
-where the matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ are defined as
+Here the matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ are defined as
 \begin{equation}
@@
 When $p$ is much greater than $m$, i.e., the dimension of $\Sigma$ is much larger than $\Omega$, one can apply Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to reduce the cost of inverting the $p\times p$ matrix for each $W_j^{(k)}$.
-Using the formulae $W_j^{(k)}$ can be rewritten in the equivalent form
+By using the formulae $W_j^{(k)}$ can be rewritten in the equivalent form
 \begin{equation}
@@
 \begin{equation}
-h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0\label{eq:nu1}
+h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log |\bar{W}^{(k)}|=0\label{eq:nu1}
 \end{equation}
@@
 \begin{equation}
-\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.\label{eq:nu2}
+\log|\Sigma|-\log |\bar{W}^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.\label{eq:nu2}
 \end{equation}
@@
 
+When $\Omega$ and $\Sigma$ are both unknown, the updating steps given in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood of matrix--$t$ distribution monotonically but the parameters $\Omega_k$ and $\Sigma_k$ may not converge due to the non-identifiability, matrix distributions are only defined up to a multiplication constant.
+This issue has also been mentioned in \citep{2013arXiv1309.6609G} for the estimation of matrix normal distributions.
+However, instead of fixing a single component in one of the scale matrices, we suggest setting the mean trace ratio
+\begin{equation}
+\frac{\textrm{tr}[\Omega_k]/m}{\textrm{tr}[\Sigma_k]/p}
+\end{equation}
+of $\Omega_k$ and $\Sigma_k$ to 1 with proper scaling at the end of each M-step.
+This ensures the convergence of the algorithm and also avoids possible numerical issues due to the large scale separation between $\Omega_k$ and $\Sigma_k$ during the iterations.
+
+\subsection{Parameters initialization}
+
+For EM method, the choice of the initial values for the parameters has significant effect on the convergence behavior as the algorithm does not guarantee reaching the global optimum.
+While this is less of an issue for matrix--$t$ distributions as the ML estimation is a convex optimization problem in $\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, it is still desirable to have an efficient method for selecting good initial values to reduce the number of EM iterations for reaching the required accuracy.
+
+For the location parameter $M$, the component-wise medians and spatial median \citep{Vardi15022000} with distance induced by Frobenius norm
+\begin{equation}
+d(T,T')=\sqrt{\textrm{tr}\left[(T-T')(T-T')^{\textrm{t}}\right]}
+\end{equation}
+are proper choices we found in the implementation and both can be computed with complexity $O(Nmp)$ with $N$ the number of matrices in the given data.
+
+For the rest of the parameters, we center the data by the initial values of $M$ and use the property of low dimensional projection of matrix--$t$ distributions to obtain the initial values.
+It is known that if $T\sim T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, $Tv$ is multivariate $t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega v}{\nu}\right)\Sigma$ and the same degress of freedom parameter $\nu$ for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}.
+This allows using efficient multivariate $t$ estimation routines for retrieving the initial values.
+
 \section{Conclusion}\label{sec:conclusion}
+
+In this note we derive and provide the EM formulae for MLE of matrix--$t$ distributions.
+The EM method here mainly consists of computing the weight matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ and we give some guidance on its numerical implementation.
+A way of choosing the initial values for the parameters is briefly discussed.
 

### 2017-02-02 paper.tex
Added more contents in the final discussion. Corrected some minor issues.

@@
 \begin{abstract}
-We provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss its practical implementation.
+We provide an expectation-maximization (EM) algorithm for the maximum likelihood estimation of matrix--$t$ distributions and discuss its practical implementation.
 The result can be applied to problems where the parameters of a matrix--$t$ distribution need to be obtained from data, such as the design of recommendation systems and remote sensing algorithms.
@@
 \section{Introduction}
-The matrix--$t$ distribution is a generalization of the multivariate $t$ distribution to matrix variates situations.
-Similar as other $t$ distributions, matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, matrix normal distributions, except the mixture parameter is a matrix rather than a scalar \citep{gupta1999matrix}.
+The matrix--$t$ distribution is a generalization of the multivariate $t$ distribution to a matrix variate situation.
+Similar as other $t$ distributions, the matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, the matrix normal distribution, except the mixture parameter is a matrix rather than a scalar \citep{gupta1999matrix}.
 Due to this, the distribution shares its importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785} and recommendation systems \citep{NIPS2007_3203}.
-In these problems, it is necessary to obtain the parameters of the distribution, which consist of a degrees of freedom parameter, a matrix location parameter, and two scale matrix parameters, that describe a set of given matrix--valued observations for further inference.
+
+These matrix variate models, including matrix normal distributions \citep{2013arXiv1309.6609G}, give better predictive results than the traditional PCA methods as the correlation structure among different observations can be captured in one of the scale matrix parameter and hence do not require the $i.i.d$ assumption.
+In these inferential problems, it is necessary to obtain the parameters of the underlying model.
+For matrix--$t$ distributions, the parameters consist of a degrees of freedom parameter, a matrix location parameter, and two scale matrix parameters, which describe a set of given matrix--valued observations for further inference.
 
@@
 However, optimizing the likelihood directly is a formidable task as the gradient equations are unwieldy slow to compute and solve.
-For $t$ distributions, it is well-known that one can exploit the property of scale mixture and apply expectation maximization (EM) technique to optimize the likelihood function \citep{10.2307/24305551}.
-The same technique can be also applied to matrix--$t$ distribution by viewing it a matrix-normal distribution with inverse Wishart distributed scale parameter.
+For $t$ distributions, it is well-known that one can exploit the mixture property and apply expectation maximization (EM) technique to optimize the likelihood function \citep{10.2307/24305551}.
+The same technique can be also applied to the matrix variate situation by viewing it as a matrix-normal distribution with inverse Wishart distributed scale parameter.
+When missing data presents, imputation of the missing values can be considered as a prediction problem and the relevant work for matrix--$t$ distributions can be found in \cite{NIPS2007_3203}.
 
-This note is organized as follows.
+This paper is organized as follows.
 In section \ref{sec:matrixt} we give a brief review on matrix--$t$ distributions.
-The notations and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section.
-EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief conclusion is given in section \ref{sec:conclusion}.
+Notations and preliminary matrix calculus results needed to present the derivation are presented sequentially in the section.
+EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief discussion is given in section \ref{sec:discussion}.
 
@@
 
-All matrices referred in this note are real.
-For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_+^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
+All matrices referred in this paper are real.
+For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_{++}^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
 The multivariate gamma function $\Gamma_{p}(\cdot)$ given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
@@
 
-Follow the definition in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
+Follow the definitions in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
 \begin{equation}
@@
 \end{equation}
-with $I_p$ the identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant for inverse SPD matrix $U=S^{-1}$
+with $I_p$ an identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant for inverse SPD matrix $U=S^{-1}$
 \begin{equation}
@@
 \begin{equation}
-\int_{\mathbf{S}_+^p}e^{-\textrm{tr}[S A]}|S|^{\alpha}\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}},\label{eq:intmultgammadef}
+\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}[S A]}|S|^{\alpha}\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}.\label{eq:intmultgammadef}
 \end{equation}
-where $A$ is a $p\times p$ nonsingular matrix.
+Here $A$ is a $p\times p$ nonsingular matrix.
 
-Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant, i.e., $T_{p,m}(M,c\,\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any constant $c>0$.
+Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant.
+It is easy to see from the density function \eqref{eq:matrixtpdf} that $T_{p,m}(M,c\,\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any constant $c>0$.
+This property needs to be addressed properly in ML estimation to give a convergence iterative algorithm.
 
@@
 Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
-With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method for MLE consists of two steps:
+With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method for ML estimation consists of two steps:
 
 \begin{itemize}
-\item At the $k$-th iteration, compute the conditional expectations with respect to $Z$
+\item At the $k$-th iteration, compute the conditional expectations with respect to $Z$ for each observation $x_i$
 \begin{equation}
-Q(\theta,\theta_k)=\sum_{i\in I}Q_i(\theta,\theta_k),\qquad Q_i(\theta,\theta_k)=\mathbb{E}_{\theta_k}\left[\log p_{\theta}(x_i,Z)|x_i\right].
+Q_i(\theta,\theta_k)=\mathbb{E}_{\theta_k}\left[\log p_{\theta}(x_i,Z)|x_i\right].
 \end{equation}
 
-\item Update $\theta$ by maximizing $Q(\theta,\theta_k)$, i.e., $\theta_{k+1} = \textrm{argmax}_{\theta}\,Q(\theta,\theta_k)$.
+\item Update $\theta$ by maximizing $Q(\theta,\theta_k):=\sum_{i\in I}Q_i(\theta,\theta_k)$ with respect to $\theta$, i.e.,
+\begin{equation}
+\theta_{k+1} = \textrm{argmax}_{\theta}\,Q(\theta,\theta_k).
+\end{equation}
 \end{itemize}
@@
 
-Given observed matrix data $\left\{T_j\right\}_{j\in J}$, for each observation $T_j$ the E-step of matrix--$t$ distributions entails computing the following integral at step $k$
+Given observed matrix data $\left\{T_j\right\}_{j\in J}$ for each observation $T_j$ the E-step of the matrix--$t$ distribution entails computing the following integral at step $k$
 \begin{equation}
-Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_+^p}f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS,\label{eq:expdef}
+Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_{++}^p}f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS,\label{eq:expdef}
 \end{equation}
@@
 We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log f(T_{j},S;\theta)$.
-Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the parameters in $\theta$ we have
+Eliminating a global factor $\frac{1}{2}$ and constant terms which do not depend on the parameters in $\theta$ we have
 \begin{equation}
@@
 \end{equation}
-The integrals to be computed are
+The expectation computation requires computing the following two integrals:
 \begin{equation}
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
-\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.
+\int_{\mathbf{S}_{++}^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
+\int_{\mathbf{S}_{++}^p}f(T_j,S;\theta_k) S^{-1}\,dS.
 \end{equation}
-They can be done by using equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively
+It can be done by using equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively, and this gives
 \begin{equation}
-\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
+\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
 \end{equation}
 \begin{equation}
-\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
+\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
 \end{equation}
@@
 \end{equation}
+
 As $\Sigma_k$ and $\Omega_k$ are both SPD, each matrix $W_j^{(k)}$ is also SPD.
@@
 When $p$ is much greater than $m$, i.e., the dimension of $\Sigma$ is much larger than $\Omega$, one can apply Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to reduce the cost of inverting the $p\times p$ matrix for each $W_j^{(k)}$.
-By using the formulae $W_j^{(k)}$ can be rewritten in the equivalent form
+By using the formulae $W_j^{(k)}$ can be rewritten in the following equivalent form
 \begin{equation}
@@
 \begin{equation}
-\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:Sigma}
+\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}.\label{eq:Sigma}
 \end{equation}
@@
 The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive.
-For this case we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$, which has the form
+In this situation we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$ which has the form
 \begin{equation}
@@
 
-When $\Omega$ and $\Sigma$ are both unknown, the updating steps given in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood of matrix--$t$ distribution monotonically but the parameters $\Omega_k$ and $\Sigma_k$ may not converge due to the non-identifiability, matrix distributions are only defined up to a multiplication constant.
-This issue has also been mentioned in \citep{2013arXiv1309.6609G} for the estimation of matrix normal distributions.
+When $\Omega$ and $\Sigma$ are both unknown, the updating steps given in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood of matrix--$t$ distribution monotonically but the parameters $\Omega_k$ and $\Sigma_k$ may not converge due to the non-identifiability of matrix--$t$ mentioned at the end of section \ref{sec:matrixt}.
+This issue has also been discussed in \citep{2013arXiv1309.6609G} for ML estimation of matrix normal distributions.
 However, instead of fixing a single component in one of the scale matrices, we suggest setting the mean trace ratio
@@
 \end{equation}
-of $\Omega_k$ and $\Sigma_k$ to 1 with proper scaling at the end of each M-step.
-This ensures the convergence of the algorithm and also avoids possible numerical issues due to the large scale separation between $\Omega_k$ and $\Sigma_k$ during the iterations.
+of $\Omega_k$ and $\Sigma_k$ to 1 by proper scaling at the end of each M-step.
+This ensures the convergence of the algorithm and also avoids possible numerical issues due to the possible large scale separation between $\Omega_k$ and $\Sigma_k$ during the iterations.
 
@@
 
-For EM method, the choice of the initial values for the parameters has significant effect on the convergence behavior as the algorithm does not guarantee reaching the global optimum.
-While this is less of an issue for matrix--$t$ distributions as the ML estimation is a convex optimization problem in $\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, it is still desirable to have an efficient method for selecting good initial values to reduce the number of EM iterations for reaching the required accuracy.
+For EM method, the initial values for the parameters has significant effect on the convergence behavior as the algorithm does not guarantee reaching the global optimum.
+While this is less of an issue for matrix--$t$ distributions as the ML estimation is a convex optimization problem in parameters $\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, it is still desirable to have an efficient method for selecting good initial values to reduce the number of EM iterations for reaching the required accuracy as the classical log EM is a first order algorithm.
 
@@
 \end{equation}
-are proper choices we found in the implementation and both can be computed with complexity $O(Nmp)$ with $N$ the number of matrices in the given data.
+are proper choices we found in the implementation. Both estimators can be computed with time complexity $O(Nmp)$ with $N$ the number of matrices in the given data.
 
 For the rest of the parameters, we center the data by the initial values of $M$ and use the property of low dimensional projection of matrix--$t$ distributions to obtain the initial values.
-It is known that if $T\sim T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, $Tv$ is multivariate $t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega v}{\nu}\right)\Sigma$ and the same degress of freedom parameter $\nu$ for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}.
-This allows using efficient multivariate $t$ estimation routines for retrieving the initial values.
+It is known that if $T\sim T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, then $Tv$ is multivariate $t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega v}{\nu}\right)\Sigma$ and the same degress of freedom parameter $\nu$ for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}.
+This allows using efficient routines for estimating multivariate $t$ distributions to retrieve the initial values.
 
-\section{Conclusion}\label{sec:conclusion}
+\section{Discussion}\label{sec:discussion}
 
-In this note we derive and provide the EM formulae for MLE of matrix--$t$ distributions.
-The EM method here mainly consists of computing the weight matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ and we give some guidance on its numerical implementation.
-A way of choosing the initial values for the parameters is briefly discussed.
+In this paper we derive and provide the EM formulae for ML estimation of matrix--$t$ distributions.
+The EM algorithm here mainly consists of computing the weight matrices $W_j^{(k)}$ and we provide some guidance on its numerical implementation.
+A method of choosing the initial values for the parameters in EM is briefly discussed.
+
+In principle, the computation burden of estimating matrix--$t$ distributions can be reduced significantly with the EM algorithm discussed here.
+However, due to the lack of optimized routines which dedicate to linear algebra computations of matrix-valued vector components, the algorithm described here can be difficult to optimize in the actual implementation.
+Also, scalability with large data can be an issue as the computation involves dense matrix inversion for each observation during the EM iterations.
+It is possible to alleviate the cost by considering matrix--$t$ distributions with structured scale matrix parameters which allow fast factorization.
+The EM relations described here can still be applied but with some modifications in the M-step.
+The application of structured scale matrix parameters in matrix--$t$ inference will be future work.
 

### 2017-02-02 paper.tex
Work on wording of introduction

@@
 \begin{abstract}
-We provide an expectation-maximization (EM) algorithm for the maximum likelihood estimation of matrix--$t$ distributions and discuss its practical implementation.
-The result can be applied to problems where the parameters of a matrix--$t$ distribution need to be obtained from data, such as the design of recommendation systems and remote sensing algorithms.
+We provide an expectation-maximization (EM) algorithm for the maximum
+likelihood estimation of matrix--$t$ distributions and discuss its
+practical implementation. The result can be applied to problems where
+the parameters of a matrix--$t$ distribution need to be obtained from
+data, such as the design of recommendation systems and remote sensing
+algorithms.
 \end{abstract}
@@
 \section{Introduction}
-The matrix--$t$ distribution is a generalization of the multivariate $t$ distribution to a matrix variate situation.
-Similar as other $t$ distributions, the matrix--$t$ distribution is a scale mixture of a particular class of multivariate normal models, the matrix normal distribution, except the mixture parameter is a matrix rather than a scalar \citep{gupta1999matrix}.
-Due to this, the distribution shares its importance in robust estimation and has been applied to many predictive tasks such as spatial interpolation on the prediction of pollution concentration \citep{KIBRIA2006785} and recommendation systems \citep{NIPS2007_3203}.
+The matrix--$t$ distribution is a generalization of the multivariate
+$t$ distribution to matrix variate domains. Similar to other $t$
+distributions, the matrix--$t$ distribution is a scale mixture of
+a particular class of multivariate normal models: In particular, it is
+a mixture of a matrix normal distribution, with a matrix-valued scale
+parameter \citep{gupta1999matrix}.  Because of its mixture nature, the
+matrix--$t$ distribution can be applied to building predictive models
+with applications in tasks such as  spatial interpolation (for
+example, to predict the concentration of pollution
+\citep{KIBRIA2006785}) and recommendation systems
+\citep{NIPS2007_3203}. Matrix models can be advantageous compared to
+more traditional PCA methods because the correlation structure among
+different observations can be captured in the scale matrix parameters,
+and hence do not require the $i.i.d$ assumption
+\citep{2013arXiv1309.6609G}. In such inferential problems, it is
+required to obtain the parameters of the underlying model, given a set
+of observed data samples. For matrix--$t$ distributions, the
+parameters consist of a degrees of freedom parameter, a matrix
+location parameter, and two scale matrix parameters.
 
-These matrix variate models, including matrix normal distributions \citep{2013arXiv1309.6609G}, give better predictive results than the traditional PCA methods as the correlation structure among different observations can be captured in one of the scale matrix parameter and hence do not require the $i.i.d$ assumption.
-In these inferential problems, it is necessary to obtain the parameters of the underlying model.
-For matrix--$t$ distributions, the parameters consist of a degrees of freedom parameter, a matrix location parameter, and two scale matrix parameters, which describe a set of given matrix--valued observations for further inference.
+The method of maximum likelihood (ML) is a common approach to estimate
+the parameters of the distribution. However, for matrix--$t$
+distributions, optimizing the likelihood directly is impractical since
+the gradient equations are unwieldy. Instead, we can exploit the fact
+that general for $t$ distributions, it is well-known that one can use
+the mixture property and apply the expectation maximization (EM)
+technique to optimize the likelihood function
+\citep{10.2307/24305551}. EM can also be  applied to the particular
+case of a matrix variate domains, by viewing the matrix--$t$
+distribution as a mixture.
 
-A common approach on estimating the parameters is through the method of maximum likelihood (ML).
-However, optimizing the likelihood directly is a formidable task as the gradient equations are unwieldy slow to compute and solve.
-For $t$ distributions, it is well-known that one can exploit the mixture property and apply expectation maximization (EM) technique to optimize the likelihood function \citep{10.2307/24305551}.
-The same technique can be also applied to the matrix variate situation by viewing it as a matrix-normal distribution with inverse Wishart distributed scale parameter.
-When missing data presents, imputation of the missing values can be considered as a prediction problem and the relevant work for matrix--$t$ distributions can be found in \cite{NIPS2007_3203}.
+The related problem of determining a sample from a matrix--$t$
+distribution given some observed entries of the matrix and parameters
+of the model was studied in \cite{NIPS2007_3203}.
 
-This paper is organized as follows.
-In section \ref{sec:matrixt} we give a brief review on matrix--$t$ distributions.
-Notations and preliminary matrix calculus results needed to present the derivation are presented sequentially in the section.
-EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief discussion is given in section \ref{sec:discussion}.
+This paper is organized as follows: In section \ref{sec:matrixt} we
+give a brief review on matrix--$t$ distributions, fix the notation and
+give some matrix calculus results. In section \ref{sec:EM}, we derive
+the EM formulae and provide relevant implementation details.  A brief
+discussion is given in section \ref{sec:discussion}.
 

### 2017-02-02 paper.tex
Wording of section 2

@@
 
-All matrices referred in this paper are real.
-For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_{++}^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$.
-The multivariate gamma function $\Gamma_{p}(\cdot)$ given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
+All matrices referred in this paper are real. The matrix determinant
+is denoted as $|\cdot|$. We use the symbol $\mathbf{S}_{++}^p$ to
+indicate the set of symmetric positive definite (SPD) matrices of
+dimension $p$. The multivariate gamma function $\Gamma_{p}(\cdot)$,
+given in terms of the standard gamma function $\Gamma(\cdot)$, is
+%
 \begin{equation}
-\Gamma_{p}(x)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(x+\frac{1-j}{2}\right).\label{eq:multgammadef}
+\Gamma_{p}(x)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(x+\frac{1-j}{2}\right)
+\ .\label{eq:multgammadef}
 \end{equation}
+%
 We also define the multivariate digamma function
@@
 \begin{equation}
-\psi_{p}(x):=\frac{d}{dx}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
+\psi_{p}(x):=\frac{d}{dx}\log\Gamma_{p}(x)=%
+\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right)\ .
 \end{equation}
 
-A matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ of dimensions $p\times m$ is specified by 4 parameters: the location matrix $M$ of dimensions $p\times m$ , the row and column SPD scale matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times m$, respectively, and the degrees of freedom parameter $\nu>0$.
-A $p \times m$ random matrix $T$ is said to be matrix--$t$ distributed if given the scale parameter $S$, a SPD matrix of dimensions $p\times p$, the matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and $S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
+A $p
+\times m$ random matrix $T$ is said to be matrix--$t$ distributed if
+given the scale parameter $S$, a SPD matrix of dimensions $p\times p$,
+the matrix $T$ is matrix normal distributed $N_{p,m}(M,S,\Omega)$ and
+$S$ is itself inverse Wishart distributed $W^{-1}(\Sigma,\nu +p -1)$:
+%
 \begin{equation}
-T\,|\,M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega)\label{eq:tdef1}
+T\,|\,M,\Sigma,\Omega,\nu,S \sim N_{p,m}(M,S,\Omega) \ ,\label{eq:tdef1}
 \end{equation}
@@
 \begin{equation}
-S\,|\,\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1).\label{eq:tdef2}
+S\,|\,\Sigma,\nu \sim W^{-1}(\Sigma,\nu+p-1)\ .\label{eq:tdef2}
 \end{equation}
-The matrix normal distribution $N_{p,m}(M,S,\Omega)$ has density
+%
+Therefore, the matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$
+of dimensions $p\times m$ is specified by 4 parameters: the location
+matrix $M$ of dimensions $p\times m$ , the row and column SPD scale
+matrices $\Sigma$ and $\Omega$ of dimensions $p\times p$ and $m\times
+m$, respectively, and the degrees of freedom parameter $\nu>0$.
+
+The density of the matrix normal distribution $N_{p,m}(M,S,\Omega)$
 \begin{equation}
-\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
+\frac{1}{(2\pi)^{pm/2}|\Omega|^{p/2}|S|^{m/2}}%
+\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)
+\label{eq:ndistpdf}
 \end{equation}
-at $X$ while the inverse Wishart distribution $W^{-1}(\Sigma,n)$ with $n>p-1$ has density
+at $X$,  while the density of the inverse Wishart distribution
+$W^{-1}(\Sigma,n)$ with $n>p-1$ is
 \begin{equation}
-\frac{|\Sigma|^{\frac{n}{2}}}{2^{\frac{n p}{2}}|S|^{\frac{n+p+1}{2}}\Gamma_{p}(\frac{n}{2})}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Sigma S^{-1}\right]\right)\label{eq:wdistpdf}
+\frac{|\Sigma|^{\frac{n}{2}}}{2^{\frac{n p}{2}}%
+|S|^{\frac{n+p+1}{2}}\Gamma_{p}(\frac{n}{2})}%
+\exp\left(-\frac{1}{2}\textrm{tr}\left[\Sigma S^{-1}\right]\right)
+\label{eq:wdistpdf}
 \end{equation}
-at $S$. Here $\Gamma_{p}(\cdot)$ is the
-multivariate gamma function defined in Eq.~\eqref{eq:multgammadef}.
+at $S$. Here $\Gamma_{p}(\cdot)$ is the multivariate gamma function
+defined in Eq.~\eqref{eq:multgammadef}.
 
-Follow the definitions in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2} and the densities given above we can see the density function $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
+From the definitions in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2},
+and the densities given above, it follows that the density function
+$f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$
+distribution $T_{p,m}(M,\Sigma,\Omega,\nu)$ satisfies
 \begin{equation}
-\frac{\Gamma_{p}\left(\frac{p+\nu+m-1}{2}\right)}{\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
+\frac{\Gamma_{p}\left(\frac{p+\nu+m-1}{2}\right)}%
+{\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)%
+\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}%
 |I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-\nu}{2}}
@@
 \end{equation}
-with $I_p$ an identity matrix of dimension $p$. The marginalization over $S$ can be obtained by using the Jacobain determinant for inverse SPD matrix $U=S^{-1}$
+%
+with $I_p$ the identity matrix of dimension $p$. The marginalization
+over $S$ can be obtained by using the Jacobain determinant of an
+inverse SPD matrix $U=S^{-1}$
+%
 \begin{equation}
@@
 \begin{equation}
-\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}[S A]}|S|^{\alpha}\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}.\label{eq:intmultgammadef}
+\int_{\mathbf{S}_{++}^p}e^{-\textrm{tr}[S A]}|S|^{\alpha}\,dS=%
+\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}
+\ .\label{eq:intmultgammadef}
 \end{equation}
-Here $A$ is a $p\times p$ nonsingular matrix.
+Where $A$ is a $p\times p$ nonsingular matrix.
 
-Similar as matrix normal distributions, matrix--$t$ distributions are defined up to a multiplicative constant.
-It is easy to see from the density function \eqref{eq:matrixtpdf} that $T_{p,m}(M,c\,\Omega,\Sigma/c,\nu)$ has the same distribution as $T_{p,m}(M,\Omega,\Sigma,\nu)$ for any constant $c>0$.
-This property needs to be addressed properly in ML estimation to give a convergence iterative algorithm.
+Similar to matrix normal distributions, matrix--$t$ distributions are
+defined up to a multiplicative constant. It is easy to see from the
+density function \eqref{eq:matrixtpdf} that
+$T_{p,m}(M,c\,\Omega,\Sigma/c,\nu)$ has the same distribution as
+$T_{p,m}(M,\Omega,\Sigma,\nu)$ for any constant $c>0$. This property
+needs to be taken into account in an iterative ML estimation algorithm
+in order to be able to reach convergence.
 

### 2017-02-02 paper.tex
Mention the joint density somewhere
This is used in the next section but not mentioned explicitly
anywhere.

@@
 
-From the definitions in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2},
-and the densities given above, it follows that the density function
+The density of the matrix--$t$ distribution can be obtained by
+computing the joint density of $S$ and $T$ taken as independent
+variables $f\left(T,S; (M,\Sigma,\Omega,\nu)\right)$, as the product
+of the matrix normal and inverse Wishart densities
+Eqs.~\eqref{eq:ndistpdf} and \eqref{eq:wdistpdf}, and then integrating
+the hidden parameter $S$  over the field of SPD matrices, as follows
+from the definitions in Eqs.~\eqref{eq:tdef1} and \eqref{eq:tdef2}.
+
+The density function
 $f(T;\left\{M,\Sigma,\Omega,\nu\right\})$ of a matrix--$t$
@@
 %
-with $I_p$ the identity matrix of dimension $p$. The marginalization
+with $I_p$ the identity matrix of dimension $p$.
+
+The marginalization
 over $S$ can be obtained by using the Jacobain determinant of an
@@
 \end{equation}
-Where $A$ is a $p\times p$ nonsingular matrix.
+Where $A$ is a $p\times p$ nonsingular matrix. The result in
+eq.~\eqref{eq:matrixtpdf} then follows by direct substitution.
 

### 2017-02-02 paper.tex
Begin working on sect3
Slightly more detailed wording.

@@
 
-\section{EM formulae for matrix--$t$ distribution}\label{sec:EM}
+\section{EM formulae for the matrix--$t$ distribution}\label{sec:EM}
 
-Given a random variable $X$ with probability density (or mass) function $p_{\theta}(\cdot)$.
-Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
-With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method for ML estimation consists of two steps:
-
+We begin by explaining how the EM method can be used to determine the
+maximum likelihood parameters for some model from a set of
+observations: Consider a random variable $X$ with probability density
+(or mass) function $p_{\theta}(\cdot)$. The EM method applies if the
+density is a marginalization over some unobserved random state $Z$,
+i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$.
+Given a set of observations $\left\{x_j\right\}_{j\in J}$, the
+classical EM method for ML estimation consists of two steps:
+%
 \begin{itemize}
-\item At the $k$-th iteration, compute the conditional expectations with respect to $Z$ for each observation $x_i$
-\begin{equation}
+\item At the $k$-th iteration, compute the conditional expectations
+	with respect to $Z$ for each observation $x_j$ \begin{equation}
 Q_i(\theta,\theta_k)=\mathbb{E}_{\theta_k}\left[\log p_{\theta}(x_i,Z)|x_i\right].
@@
 
-\item Update $\theta$ by maximizing $Q(\theta,\theta_k):=\sum_{i\in I}Q_i(\theta,\theta_k)$ with respect to $\theta$, i.e.,
+\item Update $\theta$ by maximizing $Q(\theta,\theta_k):=\sum_{j\in J}Q_j(\theta,\theta_k)$ with respect to $\theta$, i.e.,
 \begin{equation}
@@
 \end{itemize}
+The algorithm is initialized by providing some initial guess for the
+parameters of interest, $\theta_0$.
+
+We now show how the EM method is applied to matrix--$t$ distributions.
 
@@
 
-Given observed matrix data $\left\{T_j\right\}_{j\in J}$ for each observation $T_j$ the E-step of the matrix--$t$ distribution entails computing the following integral at step $k$
+Given observed matrix data $\left\{T_j\right\}_{j\in J}$ (with a total
+of $|J|$ observed matrix instances), for each observation $T_j$ the
+conditional expectation at step $k$ is
+%
 \begin{equation}
-Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_{++}^p}f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS,\label{eq:expdef}
+Q_j(\theta,\theta_k)=f(T_j;\theta_k)^{-1}\int_{\mathbf{S}_{++}^p}%
+f(T_{j},S;\theta_k)\log f\left(T_{j},S;\theta\right)\,dS%
+,\label{eq:expdef}
 \end{equation}
-where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$, $\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and $f(T,S;\theta)$ is the full likelihood of the distribution with the hidden scale parameter $S$.
+where $\theta=\left\{M,\Sigma,\Omega,\nu\right\}$,
+$\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and
+$f(T,S;\theta)$ is the full likelihood of the distribution with the
+hidden scale parameter $S$.
 

### 2017-02-02 paper.tex
Review of the last part

@@
 $\theta_k=\left\{M_k,\Sigma_k,\Omega_k,\nu_k\right\}$, and
-$f(T,S;\theta)$ is the full likelihood of the distribution with the
-hidden scale parameter $S$.
+$f(T,S;\theta)$ is the full joint likelihood of the distribution with
+the hidden scale parameter $S$.
 
-We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log f(T_{j},S;\theta)$.
-Eliminating a global factor $\frac{1}{2}$ and constant terms which do not depend on the parameters in $\theta$ we have
+We can split the integral in Eq.~\eqref{eq:expdef} by the terms of
+$\log f(T_{j},S;\theta)$. Eliminating a global factor $\frac{1}{2}$
+and constant terms which do not depend on the parameters in $\theta$
+we have
 \begin{equation}
@@
 \end{equation}
-The expectation computation requires computing the following two integrals:
+The first two terms in Eq.~\ref{eq:logf} require computing the
+following integrals:
 \begin{equation}
@@
 \end{equation}
-It can be done by using equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively, and this gives
+Note that the trace operation commutes with the matrix integration in
+the for the second term.  The integrals can be computed from
+identities obtained by taking the derivative of
+Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix
+$A$, respectively:
+%
 \begin{equation}
@@
 \end{equation}
+For the rest of the terms in Eq.~\eqref{eq:logf}, the integration over
+$S$ simply cancels out the prefactor.
 
-Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
+Thus, using the equalities Eq.~\eqref{eq:intderivative1} and
+Eq.~\eqref{eq:intderivative2}, and after some rearrangements, we
+obtain the closed form expression of $Q_j(\theta,\theta_k)$, where we
+have dropped the irrelevant constant terms
+%
 \begin{equation}
@@
 \end{equation}
+Since $\Sigma_k$ and $\Omega_k$ are both SPD, each matrix $W_j^{(k)}$
+is also SPD.
 
-As $\Sigma_k$ and $\Omega_k$ are both SPD, each matrix $W_j^{(k)}$ is also SPD.
-The computation of $W_j^{(k)}$ can be done efficiently using Cholesky factorization via LAPACK routines and the BLAS routines for symmetric matrix update.
-When $p$ is much greater than $m$, i.e., the dimension of $\Sigma$ is much larger than $\Omega$, one can apply Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to reduce the cost of inverting the $p\times p$ matrix for each $W_j^{(k)}$.
-By using the formulae $W_j^{(k)}$ can be rewritten in the following equivalent form
+The main computational challenge in our algorithm is calculating the
+$W_j^{(k)}$ matrices efficiently. We employ Cholesky factorization via
+LAPACK and the BLAS routines for symmetric
+matrix update. When $p$ is much greater than $m$, i.e., the dimension
+of $\Sigma$ is much larger than $\Omega$, one can apply the
+Sherman--Morrison--Woodbury formulae \citep{doi:10.1137/1023004} to
+reduce the cost of inverting the $p\times p$ matrix for each
+$W_j^{(k)}$. By using the formulae $W_j^{(k)}$ can be written as
+%
 \begin{equation}
@@
 \end{equation}
-with $U_j=(T_j-M_k)\Omega_k^{-1/2}$ and $\Omega_k^{1/2}$ is the upper Cholesky factor of $\Omega_k$.
-Note that in Eq.~\eqref{eq:SMW} one inverts a $m\times m$ matrix for each $j$ and the inversion (or factorization) of $\Sigma_k$ only needs to be done once in each iteration.
+with $U_j=(T_j-M_k)\Omega_k^{-1/2}$ and $\Omega_k^{1/2}$ is the upper
+Cholesky factor of $\Omega_k$. Note that in Eq.~\eqref{eq:SMW} one
+inverts a $m\times m$ matrix for each $j$ and the inversion (or
+factorization) of $\Sigma_k$ only needs to be done once in each
+iteration.
 
@@
 
-To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
+To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute
+the gradient of each parameter and equate them to $0$ to obtain
+$\theta_{k+1}$. For parameter $M$ we have
 \begin{equation}
@@
 \end{equation}
-As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
+As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the
+input data weighted by $W_j^{(k)}$
 \begin{equation}
@@
 \end{equation}
-Maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$ we obtain $\Omega_{k+1}$
+Upon maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$, we obtain
+$\Omega_{k+1}$
 \begin{equation}
@@
 \begin{itemize}
-\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{(k)}}$
+\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing
+	$Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain
+	$\Sigma_{k+1}$ in terms of the inverse of the mean of
+	$W_j^{{(k)}}$
 \begin{equation}
@@
 \end{equation}
-$\Sigma_{k+1}$ depends linearly on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
+$\Sigma_{k+1}$ depends linearly on $\nu$. Due to this, we substitute
+$\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with
+respect to $\nu$. We note that
 \[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
-is a concave function of $\nu$ for all $\nu>0$.
-The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive.
-In this situation we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$ which has the form
+is a concave function of $\nu$ for all $\nu>0$. In this
+situation we obtain $\nu_{k+1}$ by solving the equation
+$\tilde{Q}'(\nu)=0$ which has the form
 \begin{equation}
@@
 
-When $\Omega$ and $\Sigma$ are both unknown, the updating steps given in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood of matrix--$t$ distribution monotonically but the parameters $\Omega_k$ and $\Sigma_k$ may not converge due to the non-identifiability of matrix--$t$ mentioned at the end of section \ref{sec:matrixt}.
-This issue has also been discussed in \citep{2013arXiv1309.6609G} for ML estimation of matrix normal distributions.
-However, instead of fixing a single component in one of the scale matrices, we suggest setting the mean trace ratio
+When $\Omega$ and $\Sigma$ are both unknown, the updating steps given
+in Eqs.~\eqref{eq:Omega} and \eqref{eq:Sigma} increases the likelihood
+of matrix--$t$ distribution monotonically but the parameters
+$\Omega_k$ and $\Sigma_k$ may not converge due to the
+non-identifiability of matrix--$t$ mentioned at the end of section
+\ref{sec:matrixt}. This issue has also been discussed in
+\citep{2013arXiv1309.6609G} for ML estimation of matrix normal
+distributions. However, instead of fixing a single component in one of
+the scale matrices, we suggest setting the mean trace ratio
 \begin{equation}
@@
 \end{equation}
-of $\Omega_k$ and $\Sigma_k$ to 1 by proper scaling at the end of each M-step.
-This ensures the convergence of the algorithm and also avoids possible numerical issues due to the possible large scale separation between $\Omega_k$ and $\Sigma_k$ during the iterations.
+of $\Omega_k$ and $\Sigma_k$ to 1 by proper scaling at the end of each
+M-step. This ensures the convergence of the algorithm and also avoids
+possible numerical issues due to the possible large scale separation
+between $\Omega_k$ and $\Sigma_k$ in the iterative steps.
 
@@
 
-For EM method, the initial values for the parameters has significant effect on the convergence behavior as the algorithm does not guarantee reaching the global optimum.
-While this is less of an issue for matrix--$t$ distributions as the ML estimation is a convex optimization problem in parameters $\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, it is still desirable to have an efficient method for selecting good initial values to reduce the number of EM iterations for reaching the required accuracy as the classical log EM is a first order algorithm.
+In the EM method, the initial values of the parameters have
+a significant effect on the convergence behavior, since the algorithm
+does not guarantee reaching the global optimum. This is less of an
+issue for matrix--$t$ distributions because the ML estimation is
+a convex optimization problem in the parameters
+$\left\{M,\Sigma^{-1},\Omega^{-1},\nu\right\}$, but it is still
+desirable to select good initial values to reduce the number of EM
+iterations necessary to reach the required accuracy as the classical
+log EM is a first order algorithm.
 
-For the location parameter $M$, the component-wise medians and spatial median \citep{Vardi15022000} with distance induced by Frobenius norm
+For the location parameter $M$, we found that the component-wise
+median, and spatial
+median \citep{Vardi15022000} with distance induced by Frobenius norm
 \begin{equation}
@@
 \end{equation}
-are proper choices we found in the implementation. Both estimators can be computed with time complexity $O(Nmp)$ with $N$ the number of matrices in the given data.
+are good choices. Both estimators can be computed with time complexity
+$O(Nmp)$ with $N=|J|$ the number of matrices in the given data.
 
-For the rest of the parameters, we center the data by the initial values of $M$ and use the property of low dimensional projection of matrix--$t$ distributions to obtain the initial values.
-It is known that if $T\sim T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, then $Tv$ is multivariate $t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega v}{\nu}\right)\Sigma$ and the same degress of freedom parameter $\nu$ for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}.
-This allows using efficient routines for estimating multivariate $t$ distributions to retrieve the initial values.
+To obtain the rest of initial the parameters, we first center the data
+by the initial values of $M$. We then  use the property of low
+dimensional projection of matrix--$t$ distributions to obtain the
+initial values.  It is known that if $T\sim
+T_{p,m}(\mathbf{0},\Sigma,\Omega,\nu)$, then $Tv$ is a multivariate
+$t$ distributed with scale parameter $\left(\frac{v^{\textrm{t}}\Omega
+v}{\nu}\right)\Sigma$ and the same degrees of freedom parameter $\nu$
+for any nonzero $m$ dimensional vector $v$ \citep{gupta1999matrix}.
+This allows using existing routines for estimating multivariate $t$
+distributions to retrieve the initial values.
 
@@
 
-In this paper we derive and provide the EM formulae for ML estimation of matrix--$t$ distributions.
-The EM algorithm here mainly consists of computing the weight matrices $W_j^{(k)}$ and we provide some guidance on its numerical implementation.
-A method of choosing the initial values for the parameters in EM is briefly discussed.
+In this paper we derive and provide the EM formulae for ML estimation
+of matrix--$t$ distributions. The EM algorithm here mainly consists of
+computing the weight matrices $W_j^{(k)}$ and we provide some guidance
+on its numerical implementation. A method of choosing the initial
+values for the parameters in EM is briefly discussed.
 
-In principle, the computation burden of estimating matrix--$t$ distributions can be reduced significantly with the EM algorithm discussed here.
-However, due to the lack of optimized routines which dedicate to linear algebra computations of matrix-valued vector components, the algorithm described here can be difficult to optimize in the actual implementation.
-Also, scalability with large data can be an issue as the computation involves dense matrix inversion for each observation during the EM iterations.
-It is possible to alleviate the cost by considering matrix--$t$ distributions with structured scale matrix parameters which allow fast factorization.
-The EM relations described here can still be applied but with some modifications in the M-step.
-The application of structured scale matrix parameters in matrix--$t$ inference will be future work.
+In principle, the computation burden of estimating matrix--$t$
+distributions can be reduced significantly with the EM algorithm
+discussed here. However, due to the lack of optimized routines which
+dedicate to linear algebra computations of matrix-valued vector
+components, the algorithm described here can be difficult to optimize
+in the actual implementation. Also, scalability with large data can be
+an issue as the computation involves dense matrix inversion for each
+observation during the EM iterations. It is possible to alleviate the
+cost by considering matrix--$t$ distributions with structured scale
+matrix parameters which allow fast factorization. The EM relations
+described here can still be applied but with some modifications in the
+M-step. The application of structured scale matrix parameters in
+matrix--$t$ inference will be future work.
 
 \section*{Acknowledgement}
-This work is not supported by any grant and has not been presented in any regional or international meetings.
+Z.K. is supported by  the  Executive  Research  Agency  (REA)  of  the
+European Commission under the Grant Agreement PITN-GA-2012-316704
+(HiggsTools).
 
